---
title: "Mais"
output: html_document
---

Loading libraries
```{r}
library(phyloseq)
library(ggplot2)
library(tidyverse)
library(vegan)
```

Load the saved `phyloseq` object, derive the practical metadata variables, and confirm that the abundance table, taxonomy table and sample metadata are present.

```{r}
load("C:/Users/mai/OneDrive/Documents/Uni/Metagenomic Analyses of Microbiomes/Projekt/diabetes_t2_D003924/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")

pylsq #Viser et overblik over phyloseq-objektet.

sample_data(pylsq) #viser meta data for alle samples
rank_names(pylsq) #Viser hvilke taxonomiske niveauer der findes.
sample_variables(pylsq) #Viser navnene på metadata-kolonnerne.
```


Calculating alpha diversity
```{r}
# Calculate Shannon diversity for all samples
shannon <- estimate_richness(
  pylsq,
  measures = "Shannon"
)

head(shannon)
```


Visualizing the alpha diversity
```{r}
plot_richness(
  pylsq,
  measures = "Shannon"
)
```

D006262 er et MeSH ID: Healthy
D003924 er et MeSH ID: Diabetes Mellitus
```{r}
plot_richness(
  pylsq,
  measures = "Shannon",
  x = "Project.ID",
  color = "Country"
) +
  geom_boxplot(
    aes(group = Project.ID),
    alpha = 0.4,
    outlier.shape = NA
  )

```

# 3. Statistical testing alpha

The built-in `plot_richness()` function is very useful for quick exploratory visualisations, but in many analyses we also want to perform formal statistical testing to evaluate whether diversity differs significantly between groups of samples.

## Combining alpha-diversity estimates with sample metadata

To do this, it is often convenient to extract the alpha-diversity values into a standard data frame and combine them with the sample metadata stored in the phyloseq object.

The code below calculates both Shannon diversity and Observed richness separately, converts the sample names into an explicit column (`sample_id`), extracts the sample metadata, and finally merges everything into a single table.

```{r}
ps_pa <- transform_sample_counts(pylsq, function(x) {ifelse(x > 0, 1, 0)})

# Calculate Shannon diversity
shannon <- estimate_richness(
  pylsq,
  measures = "Shannon"
) %>%
  rownames_to_column("sample_id")

# Calculate observed richness
observed <- estimate_richness(
  ps_pa,
  measures = "Observed"
) %>%
  rownames_to_column("sample_id")

# Extract sample metadata
sample_df <- sample_data(pylsq) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

# Combine diversity estimates and metadata
alpha_diversity <- inner_join(
  shannon,
  observed,
  by = "sample_id"
) %>%
  inner_join(
    sample_df,
    by = "sample_id"
  )
```
The resulting `alpha_diversity` data frame contains:

* one row per sample,
* alpha-diversity estimates,
* and all associated sample metadata variables.

This format is particularly useful because it allows us to use standard tidyverse and statistical functions for downstream analyses.

You can inspect the resulting table with:
```{r}
head(alpha_diversity)
```


## Testing Shannon diversity across body sites
The output includes a p-value that helps us evaluate whether alpha-diversity distributions differ across the groups being compared.

A small p-value, commonly below `0.05`, suggests evidence against the null hypothesis that all groups have similar diversity distributions, whereas a larger p-value indicates that the observed differences could plausibly arise by chance.

```{r}
kruskal_shannon <- kruskal.test(
  Shannon ~ Disease.MESH.ID,
  data = alpha_diversity
)

kruskal_shannon
```


Beta-diversity

```{r}
ps_rel <- transform_sample_counts(
  pylsq,
  function(x) x / sum(x)
)


sample_sums(ps_rel)
```



## Calculating beta-diversity distances

phyloseq provides the function `distance()` to calculate pairwise distances between samples directly from a phyloseq object.

The result is a distance matrix in which:

* small values indicate that two samples are compositionally similar,
* and large values indicate that they are more different.

In this session, we will focus on two commonly used beta-diversity metrics:

* **Bray-Curtis distance**, which is based on abundance differences between samples.
* **Jaccard distance**, which focuses on shared versus unshared taxa.

Bray-Curtis distance is one of the most widely used measures in microbial community analysis. It is sensitive to differences in taxon abundances and is therefore commonly applied to relative-abundance data.
```{r}
bray_dist <- distance(
  ps_rel,
  method = "bray"
)

#bray_dist
#as.matrix(bray_dist)[1:5, 1:5]

```

Jaccard distance is often applied to presence/absence data and focuses on whether the same taxa are present in two samples, regardless of their abundances.
```{r}
ps_pa <- transform_sample_counts(
  pylsq,
  function(x){ifelse(x > 0.01, 1, 0)}
)

sample_sums(ps_pa)
jaccard_dist <- distance(
  ps_pa,
  method = "jaccard"
)

#as.matrix(jaccard_dist)[1:3, 1:3]
```

These two distances capture different aspects of community variation and should therefore not be expected to produce identical results.

# 3. Ordination

Distance matrices are difficult to interpret directly because they contain pairwise comparisons between all samples.

Ordination methods help us summarise these relationships in a small number of dimensions so that patterns can be visualised more easily.

phyloseq provides the function `ordinate()` to compute ordinations and `plot_ordination()` to visualise them.

In this session, we will focus on two commonly used ordination methods:

* **PCoA** (Principal Coordinates Analysis).
* **NMDS** (Non-metric Multidimensional Scaling).

## PCoA based on Bray-Curtis distance

PCoA is a useful first ordination method because it attempts to represent the pairwise sample distances in a lower-dimensional Euclidean space.

The resulting plot can be used to explore whether samples from the same body site tend to cluster together and whether different body-site categories appear compositionally distinct.
```{r}
ord_bray <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = bray_dist
)

plot_ordination(
  ps_rel,
  ord_bray,
  type = "samples",
  color = "Project.ID"
)
```



## Visualising the explained variation

For PCoA, it is often useful to inspect how much of the variation is represented by the first few axes.

```{r}
plot_scree(ord_bray)
```



Vil gerne undersøge hvor meget af variationen i mikrobiomerne der kan forklares af hver variabel.
Vi undersøger om forskelle i mikrobiomet kan forklares af hvilket projekt prøven kommer fra samt om de er syge/ikke syge. 

```{r}
meta <- data.frame(sample_data(pylsq))

adonis2(
  bray_dist ~ Project.ID + Disease.MESH.ID,
  data = meta
)
```
Her ser vi at modellen forklare ca. 14.3% af variationen i mikrobiomet. P-værdien er 0,001 dvs. modellen er statistisk signifikant og mikrobiomerne variere systematisk med variablerne. 


Nu vil jeg gerne undersøge hvad de to variable kan forklare individuelt:
```{r}
adonis2(
  bray_dist ~ Project.ID + Disease.MESH.ID,
  data = meta,
  by = "margin"
)
```
Her ser vi at project ID forklare hele 11% af variationen, hvorimod sygdomsstatus kun forkalre 1,4% (dog stadig statistisk signifikant). Dette tyder på stort geografisk bias i data.

PERMANOVA analysis based on Bray–Curtis distances showed that both Project.ID and Disease.MESH.ID were significantly associated with microbiome composition. Project.ID explained the largest proportion of variation (R² = 0.113, p = 0.001), indicating substantial study- or cohort-specific effects. Disease.MESH.ID also showed a significant association with microbiome composition (R² = 0.0138, p = 0.001), although the explained variance was considerably smaller.


