---
title: "Mais"
output: html_document
---

Loading libraries
```{r}
library(phyloseq)
library(ggplot2)
```

Load the saved `phyloseq` object, derive the practical metadata variables, and confirm that the abundance table, taxonomy table and sample metadata are present.

```{r}
load("C:/Users/mai/OneDrive/Documents/Uni/Metagenomic Analyses of Microbiomes/Projekt/diabetes_t2_D003924/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")

pylsq
sample_data(pylsq)
rank_names(pylsq)
sample_variables(pylsq)
```


Calculating alpha diversity
```{r}
# Calculate Shannon diversity for all samples
shannon <- estimate_richness(
  pylsq,
  measures = "Shannon"
)

head(shannon)
```


Visualizing the alpha diversity
```{r}
plot_richness(
  pylsq,
  measures = "Shannon"
)
```

D006262 er et MeSH ID: Healthy
D003924 er et MeSH ID: Diabetes Mellitus
```{r}
plot_richness(
  pylsq,
  measures = "Shannon",
  x = "Project.ID",
  color = "Country"
) +
  geom_boxplot(alpha = 0.4)

table(sample_data(pylsq)$Disease.name)

```


Beta-diversity
# 3. Statistical testing

The built-in `plot_richness()` function is very useful for quick exploratory visualisations, but in many analyses we also want to perform formal statistical testing to evaluate whether diversity differs significantly between groups of samples.

## Combining alpha-diversity estimates with sample metadata

To do this, it is often convenient to extract the alpha-diversity values into a standard data frame and combine them with the sample metadata stored in the phyloseq object.

The code below calculates both Shannon diversity and Observed richness separately, converts the sample names into an explicit column (`sample_id`), extracts the sample metadata, and finally merges everything into a single table.
```{r}
ps_rel <- transform_sample_counts(
  pylsq,
  function(x) x / sum(x)
)


sample_sums(ps_rel)
```



## Calculating beta-diversity distances

phyloseq provides the function `distance()` to calculate pairwise distances between samples directly from a phyloseq object.

The result is a distance matrix in which:

* small values indicate that two samples are compositionally similar,
* and large values indicate that they are more different.

In this session, we will focus on two commonly used beta-diversity metrics:

* **Bray-Curtis distance**, which is based on abundance differences between samples.
* **Jaccard distance**, which focuses on shared versus unshared taxa.

Bray-Curtis distance is one of the most widely used measures in microbial community analysis. It is sensitive to differences in taxon abundances and is therefore commonly applied to relative-abundance data.
```{r}
bray_dist <- distance(
  ps_rel,
  method = "bray"
)

bray_dist
as.matrix(bray_dist)[1:5, 1:5]

```

Jaccard distance is often applied to presence/absence data and focuses on whether the same taxa are present in two samples, regardless of their abundances.
```{r}
ps_pa <- transform_sample_counts(
  pylsq,
  function(x){ifelse(x > 0.01, 1, 0)}
)

sample_sums(ps_pa)
jaccard_dist <- distance(
  ps_pa,
  method = "jaccard"
)

jaccard_dist
as.matrix(jaccard_dist)[1:5, 1:5]
```

These two distances capture different aspects of community variation and should therefore not be expected to produce identical results.

# 3. Ordination

Distance matrices are difficult to interpret directly because they contain pairwise comparisons between all samples.

Ordination methods help us summarise these relationships in a small number of dimensions so that patterns can be visualised more easily.

phyloseq provides the function `ordinate()` to compute ordinations and `plot_ordination()` to visualise them.

In this session, we will focus on two commonly used ordination methods:

* **PCoA** (Principal Coordinates Analysis).
* **NMDS** (Non-metric Multidimensional Scaling).

## PCoA based on Bray-Curtis distance

PCoA is a useful first ordination method because it attempts to represent the pairwise sample distances in a lower-dimensional Euclidean space.

The resulting plot can be used to explore whether samples from the same body site tend to cluster together and whether different body-site categories appear compositionally distinct.
```{r}
ord_bray <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = bray_dist
)

plot_ordination(
  ps_rel,
  ord_bray,
  type = "samples",
  color = "Disease.MESH.ID"
)
```



## Visualising the explained variation

For PCoA, it is often useful to inspect how much of the variation is represented by the first few axes.

```{r}
plot_scree(ord_bray)
```

