---
title: "1 - Full dataset (genus level): exploration, diversity and cohort structure"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

# 2. Load and explore
```{r load}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/D003924_phyloseq_updated_genus.RData")
ps_genus <- pylsq
rm(pylsq)

ps_genus
```

```{r explore}
rank_names(ps_genus)
sample_variables(ps_genus)
head(sample_data(ps_genus))
```

# 3. Balance
```{r balance-disease}
table(sample_data(ps_genus)$Disease.MESH.ID, useNA = "ifany")
```
```{r balance-project-country}
table(sample_data(ps_genus)$Project.ID, useNA = "ifany")
table(sample_data(ps_genus)$Country, useNA = "ifany")
```

```{r balance-crosstabs}
table(sample_data(ps_genus)$Project.ID, sample_data(ps_genus)$Disease.MESH.ID)
table(sample_data(ps_genus)$Country, sample_data(ps_genus)$Disease.MESH.ID)
```

```{r alpha-table}
alpha <- estimate_richness(ps_genus, measures = c("Shannon")) %>%
  rownames_to_column("sample_id")

meta <- sample_data(ps_genus) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

alpha_div <- inner_join(alpha, meta, by = "sample_id") %>%
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

head(alpha_div)
```


```{r alpha-plot, fig.width=4, fig.height=4}
alpha_div %>%
  pivot_longer(c(Shannon), names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Alpha diversity",
         title = "Alpha diversity by disease status ")
```

```{r alpha-wilcox}
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
```

```{r beta-dist}
ps_rel    <- transform_sample_counts(ps_genus, function(x) x / sum(x))
bray_dist <- distance(ps_rel, method = "bray")
ord_bray  <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)

sample_df <- sample_data(ps_rel) %>% data.frame()
```

```{r pcoa-disease, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("D006262" = "#56B4E9", "D003924" = "#E74C3C"),
                     labels = c("D006262" = "Healthy", "D003924" = "T2D"),
                     name   = "Disease") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA - disease (full data, genus)")
```

```{r pcoa-project, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Project.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("PRJEB1786"   = "orange",
                                "PRJNA299502" = "forestgreen",
                                "PRJNA422434" = "purple"),
                     name = "Project") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA - project (full data, genus)")
```

```{r permanova}
adonis2(bray_dist ~ Disease.MESH.ID, data = sample_df, permutations = 999)
adonis2(bray_dist ~ Project.ID,      data = sample_df, permutations = 999)
```

```{r betadisper}
disp_disease <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(disp_disease)
disp_project <- betadisper(bray_dist, sample_df$Project.ID)
anova(disp_project)
```

```{r betadisper-plots, fig.width=7, fig.height=5}
plot(disp_disease, main = "Dispersion by disease")
plot(disp_project, main = "Dispersion by project")
```

# 6. Alpha diversity by project

```{r alpha-by-project, fig.width=8, fig.height=5}
plot_richness(
  ps_genus,
  x        = "Project.ID",
  color    = "Project.ID",
  measures = "Shannon"
) +
  geom_boxplot() +
  scale_color_manual(values = c("PRJEB1786"   = "orange",
                                "PRJNA299502" = "forestgreen",
                                "PRJNA422434" = "purple"),
                     name = "Project") +
  theme_bw() +
  labs(x = "Project", y = "Alpha diversity",
       title = "Shannon diversity by project (full data, genus)")
```

```{r kruskal-project}
kruskal.test(Shannon ~ Project.ID, data = alpha_div)
```

```{r pairwise-project}
pairwise.wilcox.test(alpha_div$Shannon, alpha_div$Project.ID, p.adjust.method = "BH")
```

# Added after submission

Alpha diversity across project ids. New plot where only sick from every cohort is in the data.
```{r alpha-by-project-only-sick, fig.width=8, fig.height=5}
alpha_div %>%
  filter(Disease == "T2D") %>%
  ggplot(aes(x = Project.ID, y = Shannon, color = Project.ID)) +
    geom_boxplot() +
    scale_color_manual(values = c("PRJEB1786"   = "orange",
                                  "PRJNA299502" = "forestgreen",
                                  "PRJNA422434" = "purple"),
                       name = "Project") +
    theme_bw() +
    labs(x = "Project", y = "Alpha diversity",
         title = "Shannon diversity by project (T2D individuals only)")
```

Testing whether the cohorts still differ in alpha diversity when only T2D individuals are kept.
```{r kruskal-project-only-sick}
sick <- alpha_div %>% filter(Disease == "T2D")
kruskal.test(Shannon ~ Project.ID, data = sick)
pairwise.wilcox.test(sick$Shannon, sick$Project.ID, p.adjust.method = "BH")
```

Same thing but for beta diversity: Bray–Curtis PCoA recomputed on T2D individuals only, coloured by project.
```{r beta-dist-only-sick}
ps_sick     <- subset_samples(ps_genus, Disease.MESH.ID == "D003924")
ps_sick_rel <- transform_sample_counts(ps_sick, function(x) x / sum(x))
bray_sick   <- distance(ps_sick_rel, method = "bray")
ord_sick    <- ordinate(ps_sick_rel, method = "PCoA", distance = bray_sick)
sample_sick <- sample_data(ps_sick_rel) %>% data.frame()
```

```{r pcoa-project-only-sick, fig.width=8, fig.height=5}
plot_ordination(ps_sick_rel, ord_sick, color = "Project.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("PRJEB1786"   = "orange",
                                "PRJNA299502" = "forestgreen",
                                "PRJNA422434" = "purple"),
                     name = "Project") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA - project (T2D individuals only)")
```

Testing whether the cohorts still differ in composition when only T2D individuals are kept.
```{r permanova-project-only-sick}
adonis2(bray_sick ~ Project.ID, data = sample_sick, permutations = 999)
disp_sick <- betadisper(bray_sick, sample_sick$Project.ID)
anova(disp_sick)
```
```{r}
plot(disp_sick, main = "Dispersion by disease")
```

