---
title: "1 — Full dataset (genus level): exploration, diversity and cohort structure"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

# 2. Load and explore

Historietid:
Firstly, we load the phyloseq object on genus level to make som preliminary exploration...
```{r load}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/D003924_phyloseq_updated_genus.RData")
ps_genus <- pylsq
rm(pylsq)

ps_genus
```

From the above we see that the phyloseq object (genus level) contains 432 samples, 58 taxa...

Now we check other information and metadata.
```{r explore}
rank_names(ps_genus)
sample_variables(ps_genus)
head(sample_data(ps_genus))
```

# 3. Balance

Then, is the data balanced between healthy/sick?
```{r balance-disease}
table(sample_data(ps_genus)$Disease.MESH.ID, useNA = "ifany")
```
Somewhat balanced. 

Then, is the data balanced across ProjectID and Country?
```{r balance-project-country}
table(sample_data(ps_genus)$Project.ID, useNA = "ifany")
table(sample_data(ps_genus)$Country,    useNA = "ifany")
```
No, not at all. Very few american samples, a few european samples and a lot of chinese samples.

Then checking healthy/sick across the different ProjectIDs and Countries.
```{r balance-crosstabs}
table(sample_data(ps_genus)$Project.ID, sample_data(ps_genus)$Disease.MESH.ID)
table(sample_data(ps_genus)$Country,    sample_data(ps_genus)$Disease.MESH.ID)
```
Not balanced at all. Only chinese healthy control???

# 4. Alpha diversity by disease

Lets investigate alpha diversity by meshid (healthy/sick). Both Shannon and Simpson. 
*We should elaborate on why these are chosen!*
```{r alpha-table}
alpha <- estimate_richness(ps_genus, measures = c("Shannon", "Simpson")) |>
  rownames_to_column("sample_id")

meta <- sample_data(ps_genus) |>
  data.frame() |>
  rownames_to_column("sample_id")

alpha_div <- inner_join(alpha, meta, by = "sample_id") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

head(alpha_div)
```


Plotting the diversity
```{r alpha-plot, fig.width=8, fig.height=4}
alpha_div |>
  pivot_longer(c(Shannon, Simpson), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Alpha diversity",
         title = "Alpha diversity by disease status (full data, genus level)")
```
A bit unexpected... T2D has higher alpha diversity than healthy? Let's investigate by doing a wilcoxon rank sum test to see if the difference in alpha diversity is significant...
```{r alpha-wilcox}
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(Simpson ~ Disease.MESH.ID, data = alpha_div)
```
Significant for both shannon and simpson...



# 5. Beta diversity by disease


Now investigating beta-diversity - Bray curtis
*Again we should explain why we chose Bray Curtis!*
```{r beta-dist}
ps_rel    <- transform_sample_counts(ps_genus, function(x) x / sum(x))
bray_dist <- distance(ps_rel, method = "bray")
ord_bray  <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)

sample_df <- sample_data(ps_rel) |> data.frame()
```

Plotting by PCoA
*NMDS as well?*
```{r pcoa-disease, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("D006262" = "#56B4E9", "D003924" = "#E74C3C"),
                     labels = c("D006262" = "Healthy", "D003924" = "T2D"),
                     name   = "Disease") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA — disease (full data, genus)")
```
Somewhat interesting at first glance. A small T2D cluster in lower left corner. Otherwise pretty mixed.
We suspect this might be to due to the data imbalance introduced by the ProjectIDs.
Let's investigate that by PCoA of Bray Curtis colored by ProjectID. 
```{r pcoa-project, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Project.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("PRJEB1786"   = "orange",
                                "PRJNA299502" = "forestgreen",
                                "PRJNA422434" = "purple"),
                     name = "Project") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA — project (full data, genus)")
```
This shows a pretty clear pattern. The ProjectIDs are different - but are they significantly different?
And also, the T2D cluster in the lower left corner was just an artefact of the two projects (non-chinese) having no healthy controls. 

Permanova followed by Betadispersion tests...
```{r permanova}
adonis2(bray_dist ~ Disease.MESH.ID, data = sample_df, permutations = 999)
adonis2(bray_dist ~ Project.ID,      data = sample_df, permutations = 999)
```
Permanova "on" (don't know if that is the wording) Disease.MESH.ID shows that Disease.MESH.ID explains roughly 3% of the total variance. The significant p-value tells us that there is a significant difference between the Disease.MESH.IDs. Also high f-statistic of 13.134. 

Permanova "on" (don't know if that is the wording) Project.ID shows that Project.ID explains roughly 13% of the total variance. -> much more than Disease.MESH.ID. The significant p-value tells us that there is a significant difference between the ProjectIDs. Also very high f-statistic of 31.732. 

To test if the difference is due to unequal dispersion we run a betadisper:
```{r betadisper}
disp_disease <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(disp_disease)
disp_project <- betadisper(bray_dist, sample_df$Project.ID)
anova(disp_project)
```
The p-value for dispersion in regards to Disease.MESH.ID is very low. This shift is mostly just due to unequal dispersion between Disease.MESH.IDs. (also clearly visible in the plot below.)

The p-value for dispersion in regards to Project.ID is still significant but not as highly significant as for Disease.MESH.ID. This suggests that differences between projects are less influenced (though influenced) by unequal dispersion and more likely reflect true difference between Project.IDs. This is again supported by the permanova result: Project.ID (R² = 0.129) compared to Disease.MESH.ID (R² = 0.030). Also, the Project.IDs are different cohorts. Very likely to have not only biological variance due to genetic/geographic/cultural differences, but also technical variance due to the samples being gathered and processed by entirely different labs. Mic drop.

Very pretty plots supporting the above.
```{r betadisper-plots, fig.width=7, fig.height=5}
plot(disp_disease, main = "Dispersion by disease")
plot(disp_project, main = "Dispersion by project")
```

# 6. Alpha diversity by project

Now we check alpha diversity (shannon) by different ProjectIDs.
```{r alpha-by-project, fig.width=8, fig.height=5}
plot_richness(
  ps_genus,
  x        = "Project.ID",
  color    = "Project.ID",
  measures = "Shannon"
) +
  geom_boxplot() +
  scale_color_manual(values = c("PRJEB1786"   = "orange",
                                "PRJNA299502" = "forestgreen",
                                "PRJNA422434" = "purple"),
                     name = "Project") +
  theme_bw() +
  labs(x = "Project", y = "Shannon",
       title = "Shannon diversity by project (full data, genus)")
```
This also might explain why T2D seems to have significantly higher alpha diversity! 
Only healthy people from china and samples from china (PRJNA422434) has lower alpha diversity than samples from the other ProjectIDs.

Testing this difference
*Why kruskals?*
```{r kruskal-project}
kruskal.test(Shannon ~ Project.ID, data = alpha_div)
```
Significant difference. Kruskal-Wallis only tells us the projects differ overall, so we follow up with pairwise Wilcoxon (BH-corrected) to see which project pairs differ.
```{r pairwise-project}
pairwise.wilcox.test(alpha_div$Shannon, alpha_div$Project.ID, p.adjust.method = "BH")
```

*Final summary of why we choose to filter*