---
title: "2 — Chinese cohort (genus level): the clean case–control comparison"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

# 2. Load and filter to China

From the previous file, we have provided arguments as to why we only will look at the chinese cohort. Therefore loading the object and instantly filtering so we only have the chinese cohort data.
```{r load-filter}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/D003924_phyloseq_updated_genus.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
```
Now the above shows that we have 359 samples and still 58 taxa on genus level.
We can also observe a more even distribution of healthy/sick.


```{r alpha-prep}
alpha <- estimate_richness(ps, measures = c("Shannon", "Simpson")) |>
  rownames_to_column("sample_id")

meta <- sample_data(ps) |>
  data.frame() |>
  rownames_to_column("sample_id")

alpha_div <- inner_join(alpha, meta, by = "sample_id") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))
```

# 3. Metadata / covariate balance between disease groups

Before testing we check whether the disease groups are balanced on the covariates (sex, recent antibiotic use, age). Any imbalance is a potential confounder. These we can adjust for in file 3.

Firstly, sex imbalance?

```{r meta-sex}
table(alpha_div$Disease, alpha_div$Sex)
chisq.test(table(alpha_div$Disease, alpha_div$Sex))
```
Not significant. *why this test?*


Now, recent antibiotics use? 
```{r meta-antibiotics}
table(alpha_div$Disease, alpha_div$Recent.antibiotics.use, useNA = "ifany")
```
None. Perfect.

Now checking age. 
```{r meta-age}
alpha_div |>
  mutate(Host.age = as.numeric(Host.age)) |>
  group_by(Disease) |>
  summarise(median_age = median(Host.age, na.rm = TRUE),
            IQR_age    = IQR(Host.age, na.rm = TRUE),
            n          = dplyr::n(),
            .groups = "drop")

wilcox.test(as.numeric(Host.age) ~ Disease.MESH.ID, data = alpha_div)
```
Somewhat big difference in median age. 44 and 55. 
Significant result. Potential confounder.

Plotting it.
```{r meta-age-plot, fig.width=5, fig.height=4}
alpha_div |>
  mutate(Host.age = as.numeric(Host.age)) |>
  ggplot(aes(Disease, Host.age, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Host age", title = "Age by disease (Chinese cohort)")
```

Lastly BMI — note it is missing for most samples, so this is only on the complete cases.
```{r meta-bmi}
alpha_div |>
  mutate(BMI = as.numeric(BMI)) |>
  group_by(Disease) |>
  summarise(median_BMI = median(BMI, na.rm = TRUE),
            IQR_BMI    = IQR(BMI, na.rm = TRUE),
            n_with_BMI = sum(!is.na(BMI)),
            n          = dplyr::n(),
            .groups = "drop")

wilcox.test(as.numeric(BMI) ~ Disease.MESH.ID, data = alpha_div)
```
~58% of BMI values are missing, and on the complete cases BMI does not differ between groups (p = 0.34). So BMI is not a confounder here, and is only used as a sensitivity covariate (n = 151) in file 3, not in the primary model.

# 4. Sequencing depth


Does alpha diversity correlate with sequencing depth?
```{r depth-by-group}
wilcox.test(Nr..reads.sequenced ~ Disease.MESH.ID, data = alpha_div)
```
From this test we find that sequencing depth differs between the meshids. (p-value = 5.582e-05)

We then need to test if it correalates with alpha diversity.
```{r depth-vs-shannon, fig.width=7, fig.height=4}
cor.test(alpha_div$Nr..reads.sequenced, alpha_div$Shannon, method = "spearman")

ggplot(alpha_div, aes(x = Nr..reads.sequenced, y = Shannon, color = Disease)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
  theme_bw() +
  labs(x = "Reads sequenced (log10)", y = "Shannon",
       title = "Sequencing depth vs Shannon diversity (Chinese cohort)")
```
Not significantly. Good. So any difference in alpha diversity is not an artefact of sequencing depth.

# 5. Alpha diversity by disease

Now looking at same alpha diversity measures as in file 1 to check if it's different when we now only have the chineses cohort in the data.
```{r alpha-plot, fig.width=8, fig.height=4}
alpha_div |>
  pivot_longer(c(Shannon, Simpson), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Alpha diversity",
         title = "Alpha diversity by disease (Chinese cohort, genus level)")
```
Testing if still significantly different. 
```{r alpha-test}
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(Simpson ~ Disease.MESH.ID, data = alpha_div)

alpha_div |>
  group_by(Disease) |>
  summarise(median_Shannon = median(Shannon),
            median_Simpson = median(Simpson),
            .groups = "drop")
```
Still significant however less than with the full dataset. This remains surprising.

# 6. Beta diversity by disease

Now investigating beta diversity. (Bray Curtis)
```{r beta-dist}
ps_rel    <- transform_sample_counts(ps, function(x) x / sum(x))
bray_dist <- distance(ps_rel, method = "bray")
ord_bray  <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)
sample_df <- sample_data(ps_rel) |> data.frame()
```


Plotting PCoA of beta diversity colored by meshid. 
```{r pcoa-disease, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("D006262" = "#56B4E9", "D003924" = "#E74C3C"),
                     labels = c("D006262" = "Healthy", "D003924" = "T2D"),
                     name   = "Disease") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA — disease (Chinese cohort)")
```
Not a lot to conclude. Maybe a small difference, but no clean seperation.

Let's test this. Frsit permanova, then betadisper. 
```{r permanova}
adonis2(bray_dist ~ Disease.MESH.ID, data = sample_df, permutations = 999)
```

```{r betadisper}
disp_disease <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(disp_disease)
```

```{r betadisper-plot, fig.width=7, fig.height=5}
plot(disp_disease, main = "Dispersion by disease")
```
