---
title: "2 — Chinese cohort (genus level): the clean case–control comparison"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

# 2. Load and filter to China

From the previous file, we have provided arguments as to why we only will look at the chinese cohort. Therefore loading the object and instantly filtering so we only have the chinese cohort data.
```{r load-filter}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/D003924_phyloseq_updated_genus.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
```
Now the above shows that we have 359 samples and still 58 taxa on genus level.
We can also observe a more even distribution of healthy/sick.
Otherwise meta data is the same 
```{r alpha-prep}
alpha <- estimate_richness(ps, measures = c("Shannon", "Simpson")) |>
  rownames_to_column("sample_id")

meta <- sample_data(ps) |>
  data.frame() |>
  rownames_to_column("sample_id")

alpha_div <- inner_join(alpha, meta, by = "sample_id") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))
```

# 3. Sequencing depth

```{r depth-by-group}
wilcox.test(Nr..reads.sequenced ~ Disease.MESH.ID, data = alpha_div)
```

```{r depth-vs-shannon, fig.width=7, fig.height=4}
cor.test(alpha_div$Nr..reads.sequenced, alpha_div$Shannon, method = "spearman")

ggplot(alpha_div, aes(x = Nr..reads.sequenced, y = Shannon, color = Disease)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
  theme_bw() +
  labs(x = "Reads sequenced (log10)", y = "Shannon",
       title = "Sequencing depth vs Shannon diversity (Chinese cohort)")
```

# 4. Alpha diversity by disease

```{r alpha-plot, fig.width=8, fig.height=4}
alpha_div |>
  pivot_longer(c(Shannon, Simpson), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Alpha diversity",
         title = "Alpha diversity by disease (Chinese cohort, genus level)")
```

```{r alpha-test}
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(Simpson ~ Disease.MESH.ID, data = alpha_div)

alpha_div |>
  group_by(Disease) |>
  summarise(median_Shannon = median(Shannon),
            median_Simpson = median(Simpson),
            .groups = "drop")
```

# 5. Beta diversity by disease

```{r beta-dist}
ps_rel    <- transform_sample_counts(ps, function(x) x / sum(x))
bray_dist <- distance(ps_rel, method = "bray")
ord_bray  <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)
sample_df <- sample_data(ps_rel) |> data.frame()
```

```{r pcoa-disease, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("D006262" = "#56B4E9", "D003924" = "#E74C3C"),
                     labels = c("D006262" = "Healthy", "D003924" = "T2D"),
                     name   = "Disease") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA — disease (Chinese cohort)")
```

```{r permanova}
adonis2(bray_dist ~ Disease.MESH.ID, data = sample_df, permutations = 999)
```

```{r betadisper}
disp_disease <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(disp_disease)
```

```{r betadisper-plot, fig.width=7, fig.height=5}
plot(disp_disease, main = "Dispersion by disease")
```
