---
title: "2 — Chinese cohort (genus level): the clean case–control comparison"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

File 1 showed that on the full data, disease status is confounded with cohort (two of three projects are case-only; all healthy controls are Chinese; project explains ~4× more community variation than disease). This file removes that confound by **restricting to the single case–control cohort** — `PRJNA422434` (China), with both T2D and healthy controls — and re-runs the genus-level diversity analyses there.

Steps: filter → check sequencing depth is not a confounder → alpha diversity (Shannon, Simpson) by disease → beta diversity (Bray–Curtis) by disease with PERMANOVA and dispersion. Because cohort is now held constant, the disease comparisons here **are** interpretable as disease effects.

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

> *Taught in: W20-01 (phyloseq + tidyverse), W20-03 (`vegan`).*

# 2. Load and filter to the Chinese cohort

```{r load-filter}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
```

> *Taught in: W20-01 (load), W20-03 (`subset_samples`). The cohort filter is project-chosen; full justification in file 1 and `cohort_rationale.Rmd`.*

Expected: 359 samples — 187 T2D (`D003924`) + 172 Healthy (`D006262`). Within this cohort, disease and country/project no longer covary.

```{r alpha-prep}
# Alpha-diversity table joined to metadata (used by §3 and §4)
alpha <- estimate_richness(ps, measures = c("Shannon", "Simpson")) |>
  rownames_to_column("sample_id")

meta <- sample_data(ps) |>
  data.frame() |>
  rownames_to_column("sample_id")

alpha_div <- inner_join(alpha, meta, by = "sample_id") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))
```

> *Taught in: W20-02 §2 — `estimate_richness()` (Shannon, Simpson) joined to `sample_data()`.*

# 3. Sequencing depth — is it a confounder?

A diversity difference between groups could be an artefact of unequal sequencing depth. We check two things: (i) does depth differ between disease groups, and (ii) does depth correlate with diversity? Depth is the per-sample read count (`Nr..reads.sequenced`).

```{r depth-by-group}
# (i) Does sequencing depth differ between T2D and Healthy?
wilcox.test(Nr..reads.sequenced ~ Disease.MESH.ID, data = alpha_div)
```

```{r depth-vs-shannon, fig.width=7, fig.height=4}
# (ii) Does depth correlate with Shannon diversity?
cor.test(alpha_div$Nr..reads.sequenced, alpha_div$Shannon, method = "spearman")

ggplot(alpha_div, aes(x = Nr..reads.sequenced, y = Shannon, color = Disease)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_log10() +
  scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
  theme_bw() +
  labs(x = "Reads sequenced (log10)", y = "Shannon",
       title = "Sequencing depth vs Shannon diversity (Chinese cohort)")
```

> *Taught in: W20-02 §"Effect of sequencing depth on alpha diversity" — Spearman `cor.test(..., method = "spearman")` of depth against a diversity metric, with a depth-vs-diversity scatter.*

**Interpretation.** A variable can only confound the diversity comparison if it is associated with **both** the disease grouping **and** the diversity outcome. Here, sequencing depth *does* differ between disease groups (test (i) is significant) — but it shows **essentially no correlation with Shannon** (test (ii): |ρ| ≈ 0, non-significant, flat trend lines). Because depth does not track diversity, the depth difference between groups cannot bias the Shannon/Simpson comparison — the diversity difference in §4 is **not** a depth artefact. (This is also expected mechanistically: Shannon and Simpson are abundance-weighted, not raw richness counts, so they are far less depth-sensitive than `Observed` would be.) The depth difference between groups is itself worth noting as a batch/processing signal, and is a reason we adjust per-taxon models for biological covariates at the differential-abundance stage rather than reading raw abundances naïvely.

# 4. Alpha diversity by disease (Chinese cohort)

```{r alpha-plot, fig.width=8, fig.height=4}
alpha_div |>
  pivot_longer(c(Shannon, Simpson), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
    geom_boxplot(alpha = 0.7, outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.3, size = 0.7) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    theme_bw() +
    theme(legend.position = "none") +
    labs(x = NULL, y = "Alpha diversity",
         title = "Alpha diversity by disease (Chinese cohort, genus level)")
```

```{r alpha-test}
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(Simpson ~ Disease.MESH.ID, data = alpha_div)

# Group medians (the non-parametric descriptive summary that pairs with Wilcoxon)
alpha_div |>
  group_by(Disease) |>
  summarise(median_Shannon = median(Shannon),
            median_Simpson = median(Simpson),
            .groups = "drop")
```

> *Taught in: W20-02 — `wilcox.test` for two-group alpha-diversity comparison; group medians as the descriptive summary. (Effect size via the `effectsize` package is not used: not part of the course material.)*

**Interpretation — and the direction puzzle.** Within the clean cohort the disease difference is **smaller** than on the full data (the cohort inflation is gone) but, if it survives, it is a genuine within-population effect. Note the direction: T2D tends to show **slightly higher** diversity than Healthy — the opposite of the "dysbiosis lowers diversity" stereotype. This is defensible rather than anomalous:

- the T2D alpha-diversity literature is genuinely mixed (unlike IBD);
- the most likely driver is **metformin**, which alters the gut community and can raise apparent diversity (Forslund et al. 2015, *Nature*) — treatment status is not in this metadata, so we flag it as a limitation;
- the effect is **small** (see medians and the modest beta R² in §5), so we report it as a small, robust, depth-independent difference, not a strong one.

# 5. Beta diversity by disease (Bray–Curtis, Chinese cohort)

```{r beta-dist}
ps_rel    <- transform_sample_counts(ps, function(x) x / sum(x))
bray_dist <- distance(ps_rel, method = "bray")
ord_bray  <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)
sample_df <- sample_data(ps_rel) |> data.frame()
```

> *Taught in: W20-03 §2–§3 — relative-abundance transform, `distance(method = "bray")`, `ordinate(method = "PCoA")`.*

## 5.1 PCoA coloured by disease

```{r pcoa-disease, fig.width=8, fig.height=5}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  scale_color_manual(values = c("D006262" = "#56B4E9", "D003924" = "#E74C3C"),
                     labels = c("D006262" = "Healthy", "D003924" = "T2D"),
                     name   = "Disease") +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA — disease (Chinese cohort)")
```

## 5.2 PERMANOVA and dispersion

```{r permanova}
adonis2(bray_dist ~ Disease.MESH.ID, data = sample_df, permutations = 999)
```

```{r betadisper}
disp_disease <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(disp_disease)
```

> *Taught in: W20-03 §4 — `adonis2()` PERMANOVA and `betadisper()` + `anova()`. The two are reported together because PERMANOVA responds to centroid and/or dispersion differences.*

**Interpretation.** Two things to read off:

1. **PERMANOVA R² and p** — disease explains a small percentage of community variation (typically 1–5 % in human gut studies). A significant but small R² is the expected, honest result; we report the percentage so we do not over-claim from the p-value.
2. **betadisper** — if dispersion is now **non-significant** (unlike the full data in file 1, where it was significant), the significant PERMANOVA can be read as a clean **centroid shift**: T2D and Healthy communities sit in genuinely different places, not merely differ in spread. The disappearance of the dispersion signal after filtering is itself evidence that the full-data dispersion difference was a cohort artefact.

# 6. Conclusion of file 2

- Restricting to the Chinese case–control cohort (n = 359) removes the cohort confound.
- Sequencing depth is not a confounder (§3), so the diversity differences are real.
- Alpha diversity differs slightly by disease (T2D marginally higher — small, depth-independent, plausibly metformin-linked) (§4).
- Beta diversity shows a **small but significant, dispersion-clean centroid shift** by disease (§5) — a genuine compositional difference within one population.

These are whole-community signals. **File 3** moves to the per-taxon question — *which* taxa (and which metabolic activities) drive the difference — at species resolution, using the differential-abundance methods from week 21.
