---
title: "3 — Chinese cohort (species level): differential abundance"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Files 1–2 worked at **genus** level and established a small, dispersion-clean disease effect within the Chinese cohort. This file moves to **species** resolution to ask the per-taxon question: *which taxa differ between T2D and Healthy?* It applies the week-21 differential-abundance toolkit — Wilcoxon, log-linear GLM, and compositional (CLR) models — each with the covariate adjustments, then compares the methods and ends with an unconstrained ordination (PCA) of the CLR profiles. Genus level is used occasionally (e.g. the top-genera composition plot) for context.

> **Note on taxonomy.** This uses the **updated** phyloseq object, in which the rank columns are correct (`Genus` = genus, `Species` = species binomial). 172 of 173 taxa carry a true binomial, so this really is species resolution. Species that the profiler could not resolve to the species level are labelled at genus; where two display labels coincide they are disambiguated with `make.unique()`.

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggh4x)
library(broom)
```

> *Taught in: W21 reference Rmd §1 (phyloseq + tidyverse + broom for per-taxon model fitting).*

# 2. Load species data and filter to China

```{r load-filter}
load("project/data/D003924_Diabetes_Mellitus,_T/species/D003924_phyloseq_updated_species.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
ntaxa(ps)
```

> *Taught in: W21 reference §1 (`load`, `ntaxa`, group table); `subset_samples` from W20-03. Cohort rationale: files 1–2.*

Expected: 359 samples (187 T2D + 172 Healthy), 173 species.

# 3. Relative-abundance long table

```{r build-rel}
ps_rel   <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel  <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel  <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()

dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel  %>% rownames_to_column("taxon"),     by = "taxon")
```

> *Taught in: W21 reference §2 — relative-abundance transform and the long `dat_rel` table joining abundance + metadata + taxonomy.*

# 4. Community composition — top 10 genera and species

Following the group's composition approach (Emilie's `Diff_21.Rmd`): collapse to genus, take the top 10 genera, merge samples into the two disease groups, and plot mean composition; then the same for the top 10 species.

```{r composition-genus, fig.width=8, fig.height=5}
ps_genus10 <- tax_glom(ps_rel, taxrank = "Genus")
top10_g    <- names(sort(taxa_sums(ps_genus10), decreasing = TRUE))[1:10]
ps_g_top   <- prune_taxa(top10_g, ps_genus10)
ps_g_disease <- merge_samples(ps_g_top, group = "Disease.MESH.ID") |>
  transform_sample_counts(function(x) x / sum(x))

plot_bar(ps_g_disease, fill = "Genus") +
  theme_bw() +
  labs(title = "Top 10 genera by disease group")
```

```{r composition-species, fig.width=8, fig.height=5}
top10_s     <- names(sort(taxa_sums(ps_rel), decreasing = TRUE))[1:10]
ps_s_top    <- prune_taxa(top10_s, ps_rel)
ps_s_disease <- merge_samples(ps_s_top, group = "Disease.MESH.ID") |>
  transform_sample_counts(function(x) x / sum(x))

plot_bar(ps_s_disease, fill = "Species") +
  theme_bw() +
  labs(title = "Top 10 species by disease group")
```

> *Taught in: W21 reference §2 (`plot_bar` of community composition); `tax_glom` to genus from W20-01; the top-N + `merge_samples` by group presentation follows the group's `Diff_21.Rmd`.*

# 5. Diagnose the data before testing

Microbiome relative abundances are zero-inflated and right-skewed; these two diagnostics justify non-parametric / log / compositional methods over a plain t-test.

```{r zero-inflation, fig.width=7, fig.height=4}
dat_rel %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(prop_zero)) +
    geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
    labs(x = "Proportion of samples with abundance = 0", y = "Number of species",
         title = "Zero-inflation across species") +
    theme_bw()
```

```{r right-skew, fig.width=9, fig.height=5}
set.seed(42)
example_taxa <- dat_rel %>%
  group_by(taxon, Species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(taxon %in% example_taxa, rel_abund > 0) %>%
  ggplot(aes(rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ Species, scales = "free") +
    labs(x = "Relative abundance (non-zero values)", y = "Count",
         title = "Right-skewed abundance distributions") +
    theme_bw()
```

> *Taught in: W21 reference §4 — zero-inflation histogram (`prop_zero`) and right-skew histograms on mid-prevalence taxa, motivating rank-based / log / compositional methods.*

# Univariate statistics

Direction convention throughout: **positive = T2D-enriched, negative = T2D-depleted** (the `D003924` coefficient relative to `D006262`).

## 6. Wilcoxon rank-sum test

```{r wilcox-results}
wilcox_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    p_value      = wilcox.test(rel_abund ~ Disease.MESH.ID, exact = FALSE)$p.value,
    mean_T2D     = mean(rel_abund[Disease.MESH.ID == "D003924"]),
    mean_Healthy = mean(rel_abund[Disease.MESH.ID == "D006262"]),
    log2_fc      = log2((mean_T2D + 1e-6) / (mean_Healthy + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

sum(wilcox_results$q_value < 0.05, na.rm = TRUE)   # n significant
```

> *Taught in: W21 reference §5 — one `wilcox.test(... ~ group, exact = FALSE)` per taxon, group means, `log2_fc`, BH-adjusted `q_value`. Wilcoxon over t-test: Alberdi reading §3; BH/FDR: §6.*

### Volcano — Wilcoxon

```{r wilcox-volcano, fig.width=8, fig.height=5}
wilcox_results %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & log2_fc > 0 ~ "T2D-enriched",
                               q_value < 0.05 & log2_fc < 0 ~ "T2D-depleted",
                               TRUE                         ~ "n.s.")) %>%
  ggplot(aes(log2_fc, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "n.s." = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "log2 fold change (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "Wilcoxon: species-level disease effect")
```

### Bar chart — top Wilcoxon hits by q-value

```{r wilcox-bar, fig.width=8, fig.height=5}
bind_rows(
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(log2_fc)  %>% head(10),
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(-log2_fc) %>% head(10)
) %>%
  distinct() %>%
  mutate(label = make.unique(as.character(Species), sep = "_"),
         direction = ifelse(log2_fc > 0, "T2D-enriched", "T2D-depleted")) %>%
  ggplot(aes(reorder(label, -q_value), log2_fc, fill = direction)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "log2 fold change (T2D vs Healthy)", fill = NULL,
         title = "Top significant species (Wilcoxon, q < 0.05)") +
    theme_bw()
```

> *Taught in: W21 reference §5 — volcano (`log2_fc` vs `-log10(q)`) and the q-ordered bar chart of significant taxa.*

## 7. Log-transformed linear models (GLM)

We fit `glm(log10(rel_abund + 1e-6) ~ ...)` per taxon and, as in week 21, build up the covariate set: **unadjusted → + Sex → + Age → + Sex + Age**. A small helper runs the per-taxon model for a given right-hand side (identical to the reference's per-taxon `group_modify`, reused across the four adjustments).

```{r glm-helper}
fit_glm <- function(rhs, label) {
  dat_rel %>%
    mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
           Sex             = factor(Sex),
           Host.age        = as.numeric(Host.age)) %>%
    group_by(taxon, Genus, Species) %>%
    group_modify(~ broom::tidy(
      glm(as.formula(paste0("log10(rel_abund + 1e-6) ~ ", rhs)),
          data = .x, family = gaussian()))) %>%
    ungroup() %>%
    filter(term == "Disease.MESH.IDD003924") %>%
    mutate(q_value = p.adjust(p.value, method = "BH"), model = label)
}

glm_all <- bind_rows(
  fit_glm("Disease.MESH.ID",                       "Unadjusted"),
  fit_glm("Disease.MESH.ID + Sex",                 "Adj. Sex"),
  fit_glm("Disease.MESH.ID + Host.age",            "Adj. Age"),
  fit_glm("Disease.MESH.ID + Sex + Host.age",      "Adj. Sex + Age")
) %>%
  mutate(model = factor(model, levels = c("Unadjusted", "Adj. Sex", "Adj. Age", "Adj. Sex + Age")))

# significant taxa per model
glm_all %>% group_by(model) %>% summarise(n_sig = sum(q_value < 0.05, na.rm = TRUE), .groups = "drop")
```

> *Taught in: W21 reference §6 (unadjusted log-GLM) and §7 (adjust the model for a covariate). Logs + pseudocount: Alberdi reading §4; linear models & covariates: §5; BH/FDR: §6.*

### Volcano — GLM across the four adjustments

```{r glm-volcano, fig.width=9, fig.height=7}
glm_all %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "n.s.")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.6) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "n.s." = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ model) +
    theme_bw() +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "GLM: disease effect across covariate adjustments")
```

### Bar chart — primary GLM (Sex + Age)

```{r glm-bar, fig.width=8, fig.height=5}
glm_primary <- glm_all %>% filter(model == "Adj. Sex + Age")

bind_rows(
  glm_primary %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(10),
  glm_primary %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(label = make.unique(as.character(Species), sep = "_"),
         direction = ifelse(estimate > 0, "T2D-enriched", "T2D-depleted")) %>%
  ggplot(aes(reorder(label, -q_value), estimate, fill = direction)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "GLM estimate (log10 RA)", fill = NULL,
         title = "Top significant species (GLM adjusted for Sex + Age)") +
    theme_bw()
```

> *Taught in: W21 reference §6–§7 (GLM volcano and q-ordered bar chart). Faceting the four adjustments into one volcano is a presentation choice to compare models compactly.*

## 8. Compositional analysis (CLR)

CLR is the compositionally-correct approach: transform each sample with the centred-log-ratio, then fit the same four models with `lm(clr ~ ...)`.

```{r clr-build}
ps_clr <- transform_sample_counts(ps, function(x) { x <- x + 1e-9; log(x) - mean(log(x)) })
otu_clr  <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr  <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()

dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr  %>% rownames_to_column("taxon"),     by = "taxon")
```

```{r clr-helper}
fit_clr <- function(rhs, label) {
  dat_clr %>%
    mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
           Sex             = factor(Sex),
           Host.age        = as.numeric(Host.age)) %>%
    group_by(taxon, Genus, Species) %>%
    group_modify(~ broom::tidy(
      lm(as.formula(paste0("clr ~ ", rhs)), data = .x))) %>%
    ungroup() %>%
    filter(term == "Disease.MESH.IDD003924") %>%
    mutate(q_value = p.adjust(p.value, method = "BH"), model = label)
}

clr_all <- bind_rows(
  fit_clr("Disease.MESH.ID",                   "Unadjusted"),
  fit_clr("Disease.MESH.ID + Sex",             "Adj. Sex"),
  fit_clr("Disease.MESH.ID + Host.age",        "Adj. Age"),
  fit_clr("Disease.MESH.ID + Sex + Host.age",  "Adj. Sex + Age")
) %>%
  mutate(model = factor(model, levels = c("Unadjusted", "Adj. Sex", "Adj. Age", "Adj. Sex + Age")))

clr_all %>% group_by(model) %>% summarise(n_sig = sum(q_value < 0.05, na.rm = TRUE), .groups = "drop")
```

> *Taught in: W21 reference §9 — CLR transform (pseudocount + `log(x) - mean(log(x))`) and `lm(clr ~ group + covariates)` per taxon. Compositionality: Alberdi reading §7.*

### Volcano — CLR across the four adjustments

```{r clr-volcano, fig.width=9, fig.height=7}
clr_all %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "n.s.")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.6) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "n.s." = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    facet_wrap(~ model) +
    theme_bw() +
    labs(x = "CLR estimate (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "CLR: disease effect across covariate adjustments")
```

### Bar chart — primary CLR (Sex + Age)

```{r clr-bar, fig.width=8, fig.height=5}
clr_primary <- clr_all %>% filter(model == "Adj. Sex + Age")

bind_rows(
  clr_primary %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(10),
  clr_primary %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(label = make.unique(as.character(Species), sep = "_"),
         direction = ifelse(estimate > 0, "T2D-enriched", "T2D-depleted")) %>%
  ggplot(aes(reorder(label, -q_value), estimate, fill = direction)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "CLR estimate (T2D vs Healthy)", fill = NULL,
         title = "Top significant species (CLR adjusted for Sex + Age)") +
    theme_bw()
```

> *Taught in: W21 reference §9 (CLR volcano and q-ordered bar chart).*

# 9. Compare methods

Which species are significant across methods? We compare the three headline analyses — Wilcoxon, GLM (Sex + Age) and CLR (Sex + Age).

```{r comparison-build}
comparison <- wilcox_results %>%
  select(taxon, Genus, Species, wilcox_log2fc = log2_fc, wilcox_q = q_value) %>%
  left_join(glm_primary %>% select(taxon, glm_estimate = estimate, glm_q = q_value), by = "taxon") %>%
  left_join(clr_primary %>% select(taxon, clr_estimate = estimate, clr_q = q_value), by = "taxon")
```

### Significant species per method

```{r sig-summary}
tibble(
  method = c("Wilcoxon", "GLM (Sex+Age)", "CLR (Sex+Age)"),
  n_significant = c(sum(wilcox_results$q_value < 0.05, na.rm = TRUE),
                    sum(glm_primary$q_value     < 0.05, na.rm = TRUE),
                    sum(clr_primary$q_value     < 0.05, na.rm = TRUE))
)
```

### Species significant in all three methods

```{r consensus}
comparison %>%
  filter(wilcox_q < 0.05, glm_q < 0.05, clr_q < 0.05) %>%
  select(Genus, Species, wilcox_q, glm_q, clr_q) %>%
  arrange(clr_q)
```

### Concordance — number of methods supporting each species

```{r concordance, fig.width=6, fig.height=4}
comparison %>%
  mutate(n_methods = (wilcox_q < 0.05) + (glm_q < 0.05) + (clr_q < 0.05)) %>%
  count(n_methods) %>%
  filter(!is.na(n_methods)) %>%
  ggplot(aes(factor(n_methods), n)) +
    geom_col(fill = "steelblue") +
    labs(x = "Number of methods with q < 0.05", y = "Number of species",
         title = "Cross-method concordance") +
    theme_bw()
```

### Effect-size agreement — Wilcoxon vs CLR

```{r wilcox-vs-clr, fig.width=7, fig.height=5}
comparison %>%
  mutate(sig = case_when(wilcox_q < 0.05 & clr_q < 0.05 ~ "Both",
                         wilcox_q < 0.05                ~ "Wilcoxon only",
                         clr_q < 0.05                   ~ "CLR only",
                         TRUE                           ~ "Neither")) %>%
  ggplot(aes(wilcox_log2fc, clr_estimate, color = sig)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("Both" = "#E74C3C", "Wilcoxon only" = "#E67E22",
                                  "CLR only" = "#3498DB", "Neither" = "grey70")) +
    labs(x = "Wilcoxon log2 fold change", y = "CLR estimate", color = NULL,
         title = "Effect-size agreement: Wilcoxon vs CLR") +
    theme_bw()
```

> *Taught in: W21 reference §10 — significant-per-method counts, consensus taxa, concordance bar chart, and cross-method effect-size scatter. Same-sign points indicate the methods agree on direction.*

# Multivariate statistics

## 10. PCA of CLR-transformed species profiles

PCA on the CLR matrix is the unconstrained view of the dominant community gradients — coloured by disease to see whether disease aligns with any major axis.

```{r pca, fig.width=7, fig.height=5}
clr_mat   <- as.matrix(otu_clr)
meta_pca  <- meta_clr[rownames(clr_mat), ]
clr_pca   <- prcomp(clr_mat, center = TRUE, scale. = FALSE)
pca_var   <- round(100 * clr_pca$sdev^2 / sum(clr_pca$sdev^2), 1)

as.data.frame(clr_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_pca %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy")) %>%
  ggplot(aes(PC1, PC2, color = Disease)) +
    geom_point(size = 2, alpha = 0.8) +
    stat_ellipse(level = 0.95) +
    scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         title = "PCA of CLR-transformed species profiles") +
    theme_bw()
```

> *Taught in: W21 reference §11 — `prcomp` on the CLR matrix, PC1/PC2 coloured by group, % variance from `sdev^2`.*

# 11. Reading the results / next steps

- **Few species clear FDR**, and the headline hit is **Roseburia intestinalis (T2D-depleted)** — a butyrate producer. This *resolves the genus-level Roseburia signal to a named species*, which is the most informative kind of finding. Where genus-level signals do **not** resolve to a single species, that dilution is itself informative (the genus is the more reliable level — see week-21 genus analysis).
- The four adjustments (unadjusted → Sex → Age → Sex + Age) let us check that the disease effect is **robust to covariates**; the faceted volcanoes and the per-model significant counts show how little the hits move.
- The three methods (Wilcoxon / GLM / CLR) are compared in §9; consensus species (significant in all three) are the most robust candidates.

**Not yet decided (next file / section):** the genus↔species reconciliation table and the **metabolic-guild interpretation** (mapping the differential taxa to SCFA producers, mucin degraders, etc. — the project's headline question), plus optionally constrained ordination (RDA). These live in `synthesis.Rmd` and can be folded in once the team agrees the final scope.
