---
title: "3 — Chinese cohort (species level): differential abundance"
author: "Group 6"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(ggplot2)
library(broom)
```

# 2. Load species data and filter to China

```{r load-filter}
load("project/data/D003924_Diabetes_Mellitus,_T/species/D003924_phyloseq_updated_species.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
ntaxa(ps)
```

# 3. Relative-abundance long table

```{r build-rel}
ps_rel   <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel  <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- as(sample_data(ps_rel), "matrix") %>% as.data.frame()
tax_rel  <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()

dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel  %>% rownames_to_column("taxon"),     by = "taxon")
```

# 4. Community composition

```{r composition-genus, fig.width=8, fig.height=5}
ps_genus10 <- tax_glom(ps_rel, taxrank = "Genus")
top10_g    <- names(sort(taxa_sums(ps_genus10), decreasing = TRUE))[1:10]
ps_g_top   <- prune_taxa(top10_g, ps_genus10)
ps_g_disease <- merge_samples(ps_g_top, group = "Disease.MESH.ID") |>
  transform_sample_counts(function(x) x / sum(x))

plot_bar(ps_g_disease, fill = "Genus") +
  scale_x_discrete(labels = c("D006262" = "Healthy", "D003924" = "T2D")) +
  theme_bw() +
  labs(x = "Disease group", y = "Relative abundance",
       title = "Top 10 genera by disease group")
```

```{r composition-species, fig.width=8, fig.height=5}
top10_s      <- names(sort(taxa_sums(ps_rel), decreasing = TRUE))[1:10]
ps_s_top     <- prune_taxa(top10_s, ps_rel)
ps_s_disease <- merge_samples(ps_s_top, group = "Disease.MESH.ID") |>
  transform_sample_counts(function(x) x / sum(x))

plot_bar(ps_s_disease, fill = "Species") +
  scale_x_discrete(labels = c("D006262" = "Healthy", "D003924" = "T2D")) +
  theme_bw() +
  labs(x = "Disease group", y = "Relative abundance",
       title = "Top 10 species by disease group")
```

# 5. Diagnostics

```{r zero-inflation, fig.width=7, fig.height=4}
dat_rel %>%
  group_by(taxon) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(prop_zero)) +
    geom_histogram(binwidth = 0.05, fill = "steelblue", color = "white") +
    labs(x = "Proportion of samples with abundance = 0", y = "Number of species",
         title = "Zero-inflation across species") +
    theme_bw()
```

```{r right-skew, fig.width=9, fig.height=5}
set.seed(42)
example_taxa <- dat_rel %>%
  group_by(taxon, Species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(taxon %in% example_taxa, rel_abund > 0) %>%
  ggplot(aes(rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ Species, scales = "free") +
    labs(x = "Relative abundance (non-zero values)", y = "Count",
         title = "Right-skewed abundance distributions") +
    theme_bw()
```

# 6. Wilcoxon rank-sum test

```{r wilcox-results}
wilcox_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    p_value      = wilcox.test(rel_abund ~ Disease.MESH.ID, exact = FALSE)$p.value,
    mean_T2D     = mean(rel_abund[Disease.MESH.ID == "D003924"]),
    mean_Healthy = mean(rel_abund[Disease.MESH.ID == "D006262"]),
    log2_fc      = log2((mean_T2D + 1e-6) / (mean_Healthy + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

wilcox_results %>% arrange(q_value) %>% head(10)
```

```{r wilcox-volcano, fig.width=8, fig.height=5}
wilcox_results %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & log2_fc > 0 ~ "T2D-enriched",
                               q_value < 0.05 & log2_fc < 0 ~ "T2D-depleted",
                               TRUE                         ~ "Not significant")) %>%
  ggplot(aes(log2_fc, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "log2 fold change (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "Wilcoxon: Differential abundance between Disease.MESH.ID groups")
```

```{r wilcox-bar, fig.width=8, fig.height=5}
bind_rows(
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(log2_fc)  %>% head(10),
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(-log2_fc) %>% head(10)
) %>%
  distinct() %>%
  mutate(label = make.unique(as.character(Species), sep = "_"),
         direction = ifelse(log2_fc > 0, "T2D-enriched", "T2D-depleted")) %>%
  ggplot(aes(reorder(label, -q_value), log2_fc, fill = direction)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "log2 fold change (T2D vs Healthy)", fill = NULL,
         title = "Top significant species (Wilcoxon, q < 0.05)") +
    theme_bw()
```

# 7. GLM — unadjusted

```{r glm-unadj}
glm_unadj <- dat_rel %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924"))) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID,
                                 data = .x, family = gaussian()))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

glm_unadj %>% arrange(q_value) %>% head(10)
```

```{r glm-unadj-volcano, fig.width=8, fig.height=5}
glm_unadj %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "GLM unadjusted: Differential abundance between Disease.MESH.ID groups")
```

# 8. GLM — adjusted for Sex

```{r glm-sex}
glm_sex <- dat_rel %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Sex = factor(Sex)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID + Sex,
                                 data = .x, family = gaussian()))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

glm_sex %>% arrange(q_value) %>% head(10)
```

```{r glm-sex-volcano, fig.width=8, fig.height=5}
glm_sex %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "GLM adjusted for Sex: Differential abundance between Disease.MESH.ID groups")
```

# 9. GLM — adjusted for Age

```{r glm-age}
glm_age <- dat_rel %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Host.age = as.numeric(Host.age)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID + Host.age,
                                 data = .x, family = gaussian()))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

glm_age %>% arrange(q_value) %>% head(10)
```

```{r glm-age-volcano, fig.width=8, fig.height=5}
glm_age %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "GLM adjusted for Age: Differential abundance between Disease.MESH.ID groups")
```

# 10. GLM — adjusted for Sex + Age

```{r glm-sexage}
glm_sexage <- dat_rel %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Sex = factor(Sex), Host.age = as.numeric(Host.age)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID + Sex + Host.age,
                                 data = .x, family = gaussian()))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

glm_sexage %>% arrange(q_value) %>% head(10)
```

```{r glm-sexage-volcano, fig.width=8, fig.height=5}
glm_sexage %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "GLM adjusted for Sex + Age: Differential abundance between Disease.MESH.ID groups")
```

# 11. CLR build

```{r clr-build}
ps_clr <- transform_sample_counts(ps, function(x) { x <- x + 1e-9; log(x) - mean(log(x)) })
otu_clr  <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- as(sample_data(ps_clr), "matrix") %>% as.data.frame()
tax_clr  <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()

dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr  %>% rownames_to_column("taxon"),     by = "taxon")
```

# 12. CLR — unadjusted

```{r clr-unadj}
clr_unadj <- dat_clr %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924"))) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(lm(clr ~ Disease.MESH.ID, data = .x))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

clr_unadj %>% arrange(q_value) %>% head(10)
```

```{r clr-unadj-volcano, fig.width=8, fig.height=5}
clr_unadj %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "CLR estimate (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "CLR unadjusted: Differential abundance between Disease.MESH.ID groups")
```

# 13. CLR — adjusted for Sex

```{r clr-sex}
clr_sex <- dat_clr %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Sex = factor(Sex)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(lm(clr ~ Disease.MESH.ID + Sex, data = .x))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

clr_sex %>% arrange(q_value) %>% head(10)
```

```{r clr-sex-volcano, fig.width=8, fig.height=5}
clr_sex %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "CLR estimate (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "CLR adjusted for Sex: Differential abundance between Disease.MESH.ID groups")
```

# 14. CLR — adjusted for Age

```{r clr-age}
clr_age <- dat_clr %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Host.age = as.numeric(Host.age)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(lm(clr ~ Disease.MESH.ID + Host.age, data = .x))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

clr_age %>% arrange(q_value) %>% head(10)
```

```{r clr-age-volcano, fig.width=8, fig.height=5}
clr_age %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "CLR estimate (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "CLR adjusted for Age: Differential abundance between Disease.MESH.ID groups")
```

# 15. CLR — adjusted for Sex + Age

```{r clr-sexage}
clr_sexage <- dat_clr %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
         Sex = factor(Sex), Host.age = as.numeric(Host.age)) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ broom::tidy(lm(clr ~ Disease.MESH.ID + Sex + Host.age, data = .x))) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))

clr_sexage %>% arrange(q_value) %>% head(10)
```

```{r clr-sexage-volcano, fig.width=8, fig.height=5}
clr_sexage %>%
  mutate(neg_log10_q = -log10(q_value),
         direction = case_when(q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
                               q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
                               TRUE                          ~ "Not significant")) %>%
  ggplot(aes(estimate, neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 1.8) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB", "Not significant" = "grey70")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    theme_bw() +
    labs(x = "CLR estimate (T2D vs Healthy)", y = "-log10(q)", color = NULL,
         title = "CLR adjusted for Sex + Age: Differential abundance between Disease.MESH.ID groups")
```

# 16. PCA of CLR-transformed species

```{r pca, fig.width=7, fig.height=5}
clr_mat  <- as.matrix(otu_clr)
meta_pca <- meta_clr[rownames(clr_mat), ]
clr_pca  <- prcomp(clr_mat, center = TRUE, scale. = FALSE)
pca_var  <- round(100 * clr_pca$sdev^2 / sum(clr_pca$sdev^2), 1)

as.data.frame(clr_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_pca %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy")) %>%
  ggplot(aes(PC1, PC2, color = Disease)) +
    geom_point(size = 2, alpha = 0.8) +
    stat_ellipse(level = 0.95) +
    scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         title = "PCA of CLR-transformed species profiles") +
    theme_bw()
```

# 17. Constrained ordination (RDA)

```{r rda-fit}
clr_mat  <- as.matrix(otu_clr)
meta_rda <- meta_clr[rownames(clr_mat), ]
meta_rda$Disease.MESH.ID <- factor(meta_rda$Disease.MESH.ID, levels = c("D006262", "D003924"))
meta_rda$Sex             <- factor(meta_rda$Sex)
meta_rda$Host.age        <- as.numeric(meta_rda$Host.age)

keep        <- complete.cases(meta_rda[, c("Disease.MESH.ID", "Sex", "Host.age")])
clr_mat_rda <- clr_mat[keep, ]
meta_rda    <- meta_rda[keep, ]

rda_fit <- rda(clr_mat_rda ~ Disease.MESH.ID + Sex + Host.age, data = meta_rda)
rda_fit
```

```{r rda-anova}
anova(rda_fit, permutations = 999)
anova(rda_fit, by = "term", permutations = 999)
```

```{r rda-biplot, fig.width=8, fig.height=6}
pct <- round(100 * rda_fit$CCA$eig / rda_fit$tot.chi, 1)

site <- scores(rda_fit, display = "sites") %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_rda %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

bp <- scores(rda_fit, display = "bp") %>% as.data.frame() %>% rownames_to_column("predictor")
arrow_scale <- max(abs(c(site$RDA1, site$RDA2))) * 0.8 / max(sqrt(bp$RDA1^2 + bp$RDA2^2))
bp <- bp %>% mutate(RDA1 = RDA1 * arrow_scale, RDA2 = RDA2 * arrow_scale)

ggplot(site, aes(RDA1, RDA2)) +
  geom_point(aes(color = Disease), size = 2, alpha = 0.8) +
  stat_ellipse(aes(color = Disease), level = 0.95) +
  geom_segment(data = bp, aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")), color = "black", linewidth = 0.6) +
  geom_text(data = bp, aes(x = RDA1 * 1.1, y = RDA2 * 1.1, label = predictor),
            color = "black", size = 3) +
  scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E74C3C")) +
  labs(x = paste0("RDA1 (", pct[1], "%)"), y = paste0("RDA2 (", pct[2], "%)"),
       color = "Disease",
       title = "RDA — species community constrained by disease, sex and age") +
  theme_bw()
```

# 18. Agreement across tests

```{r all-tests}
all_tests <- bind_rows(
  wilcox_results %>% transmute(taxon, Genus, Species, method = "Wilcoxon",     effect = log2_fc, q_value),
  glm_unadj      %>% transmute(taxon, Genus, Species, method = "GLM unadj",    effect = estimate, q_value),
  glm_sex        %>% transmute(taxon, Genus, Species, method = "GLM Sex",      effect = estimate, q_value),
  glm_age        %>% transmute(taxon, Genus, Species, method = "GLM Age",      effect = estimate, q_value),
  glm_sexage     %>% transmute(taxon, Genus, Species, method = "GLM Sex+Age",  effect = estimate, q_value),
  clr_unadj      %>% transmute(taxon, Genus, Species, method = "CLR unadj",    effect = estimate, q_value),
  clr_sex        %>% transmute(taxon, Genus, Species, method = "CLR Sex",      effect = estimate, q_value),
  clr_age        %>% transmute(taxon, Genus, Species, method = "CLR Age",      effect = estimate, q_value),
  clr_sexage     %>% transmute(taxon, Genus, Species, method = "CLR Sex+Age",  effect = estimate, q_value)
) %>%
  mutate(method = factor(method, levels = c(
    "Wilcoxon", "GLM unadj", "GLM Sex", "GLM Age", "GLM Sex+Age",
    "CLR unadj", "CLR Sex", "CLR Age", "CLR Sex+Age")))
```

```{r sig-summary}
all_tests %>%
  group_by(method) %>%
  summarise(n_significant = sum(q_value < 0.05, na.rm = TRUE), .groups = "drop")
```

```{r consensus}
all_tests %>%
  group_by(taxon, Genus, Species) %>%
  summarise(n_tests_sig = sum(q_value < 0.05, na.rm = TRUE), .groups = "drop") %>%
  filter(n_tests_sig > 0) %>%
  arrange(desc(n_tests_sig))
```

```{r concordance, fig.width=6, fig.height=4}
all_tests %>%
  group_by(taxon) %>%
  summarise(n_tests_sig = sum(q_value < 0.05, na.rm = TRUE), .groups = "drop") %>%
  filter(n_tests_sig > 0) %>%
  count(n_tests_sig) %>%
  ggplot(aes(factor(n_tests_sig), n)) +
    geom_col(fill = "steelblue") +
    labs(x = "Number of tests with q < 0.05 (of 9)", y = "Number of species",
         title = "How many tests agree per species") +
    theme_bw()
```

```{r agreement-heatmap, fig.width=8, fig.height=6}
sig_taxa <- all_tests %>% filter(q_value < 0.05) %>% distinct(taxon) %>% pull(taxon)

label_map <- all_tests %>%
  distinct(taxon, Species) %>%
  mutate(label = make.unique(as.character(Species), sep = "_"))

all_tests %>%
  filter(taxon %in% sig_taxa) %>%
  left_join(label_map, by = c("taxon", "Species")) %>%
  mutate(signed_neg_log10_q = sign(effect) * pmin(-log10(q_value), 10)) %>%
  ggplot(aes(method, label, fill = signed_neg_log10_q)) +
    geom_tile(color = "white") +
    geom_text(aes(label = ifelse(q_value < 0.05, "*", "")), size = 4) +
    scale_fill_gradient2(low = "#3498DB", mid = "white", high = "#E74C3C", midpoint = 0,
                         name = "signed\n-log10(q)") +
    labs(x = NULL, y = NULL,
         title = "Which taxa the tests agree on (* = q < 0.05)") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
