---
title: "Alpha diversity — filtered cohort"
author: "Anton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Alpha diversity within the filtered Chinese cohort (`Project.ID == "PRJNA422434"`, 187 T2D + 172 Healthy). All analyses run at both genus and species level. The headline test is Wilcoxon rank-sum; effect size, depth confounding, and covariate adjustment are also reported.

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(effectsize)
```

# 2. Load and filter data

Load the original phyloseq objects and restrict to the Chinese cohort (`Project.ID == "PRJNA422434"`). The rationale for this filter is documented separately; in short, the other two projects in the dataset are case-only and cannot contribute to a case-control comparison.

```{r load-filter}
# Genus level
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps_genus_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

# Species level
load("project/data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData")
ps_species_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps_genus_clean
ps_species_clean
```

# 3. Build a tidy alpha-diversity table at both levels

```{r alpha-table}
make_alpha_df <- function(ps, level_label) {
  alpha <- estimate_richness(ps, measures = c("Shannon", "Simpson", "InvSimpson")) |>
    rownames_to_column("sample_id")

  meta <- sample_data(ps) |>
    data.frame() |>
    rownames_to_column("sample_id")

  inner_join(alpha, meta, by = "sample_id") |>
    mutate(level = level_label)
}

alpha_genus   <- make_alpha_df(ps_genus_clean,   "Genus")
alpha_species <- make_alpha_df(ps_species_clean, "Species")
alpha_all     <- bind_rows(alpha_genus, alpha_species)

head(alpha_all)
```

Note: `Observed` is omitted because the OTU table contains relative abundances rather than integer counts, which is the data structure `Observed`/`Chao1`/`ACE` assume.

# 4. Wilcoxon tests per metric and level

```{r wilcoxon-tests}
metrics_to_test <- c("Shannon", "Simpson", "InvSimpson")
levels_to_test  <- c("Genus", "Species")

wilcox_results <- tibble()

for (lvl in levels_to_test) {
  sub <- alpha_all |> filter(level == lvl)
  for (m in metrics_to_test) {
    w <- wilcox.test(reformulate("Disease.MESH.ID", m), data = sub)
    wilcox_results <- bind_rows(
      wilcox_results,
      tibble(
        level     = lvl,
        metric    = m,
        W         = unname(w$statistic),
        p_value   = w$p.value,
        median_T2D = median(sub[[m]][sub$Disease.MESH.ID == "D003924"]),
        median_HC  = median(sub[[m]][sub$Disease.MESH.ID == "D006262"])
      )
    )
  }
}

wilcox_results <- wilcox_results |>
  mutate(p_value = signif(p_value, 3))

wilcox_results
```

# 5. Effect sizes (rank-biserial correlation)

```{r effect-sizes}
effect_results <- tibble()

for (lvl in levels_to_test) {
  sub <- alpha_all |> filter(level == lvl)
  for (m in metrics_to_test) {
    rb <- rank_biserial(reformulate("Disease.MESH.ID", m), data = sub)
    effect_results <- bind_rows(
      effect_results,
      tibble(
        level    = lvl,
        metric   = m,
        r        = rb$r_rank_biserial,
        CI_lower = rb$CI_low,
        CI_upper = rb$CI_high
      )
    )
  }
}

effect_results <- effect_results |>
  mutate(across(c(r, CI_lower, CI_upper), \(x) round(x, 3)))

effect_results
```

Effect-size interpretation (rank-biserial): |r| < 0.1 negligible, 0.1–0.3 small, 0.3–0.5 medium, > 0.5 large.

# 6. Depth confounding

A diversity difference could in principle reflect differential sequencing depth between groups. Spearman correlation between depth (`Nr..reads.sequenced`) and Shannon diversity tests this directly.

```{r depth-check}
depth_results <- alpha_all |>
  group_by(level) |>
  summarise(
    rho      = cor(Nr..reads.sequenced, Shannon, method = "spearman"),
    p_value  = cor.test(Nr..reads.sequenced, Shannon, method = "spearman")$p.value
  ) |>
  mutate(across(c(rho, p_value), \(x) signif(x, 3)))

depth_results
```

A small |ρ| (< 0.1) and a non-significant p-value indicate no meaningful depth confound.

# 7. Visualise alpha diversity by disease

```{r plot-alpha, fig.width=9, fig.height=6}
alpha_long <- alpha_all |>
  select(sample_id, level, Disease.MESH.ID,
         Shannon, Simpson, InvSimpson) |>
  pivot_longer(cols = c(Shannon, Simpson, InvSimpson),
               names_to = "metric", values_to = "value") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

ggplot(alpha_long, aes(x = Disease, y = value, fill = Disease)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_grid(metric ~ level, scales = "free_y") +
  scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#EE4B2B")) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Alpha diversity in the Chinese cohort",
       subtitle = "Shannon, Simpson, Inverse Simpson at genus and species level")
```

# 8. Covariate adjustment

Within the filtered cohort, additional covariates may still confound the disease effect — BMI, host age, and sex are biologically plausible candidates. Wilcoxon does not natively adjust for covariates, so we switch to a linear model on Shannon diversity. This is reasonable because Shannon at sample level is approximately continuous and not extremely skewed.

```{r covariate-check}
# Inspect covariate availability
summary(alpha_genus$BMI)
summary(alpha_genus$Host.age)
table(alpha_genus$Sex, alpha_genus$Disease.MESH.ID, useNA = "ifany")
```

If `BMI` has many NAs, those samples are dropped from the adjusted model only — the unadjusted comparison still uses all samples.

```{r adjusted-models}
alpha_genus_complete <- alpha_genus |>
  filter(!is.na(BMI), !is.na(Host.age), !is.na(Sex))

mod_unadjusted <- lm(Shannon ~ Disease.MESH.ID,
                     data = alpha_genus_complete)

mod_adjusted   <- lm(Shannon ~ Disease.MESH.ID + BMI + Host.age + Sex,
                     data = alpha_genus_complete)

summary(mod_unadjusted)$coefficients
summary(mod_adjusted)$coefficients
```

The key comparison is the `Disease.MESH.IDD006262` coefficient between the unadjusted and adjusted models. If the magnitude is similar and the term remains significant after adjustment, the disease effect is not explained by BMI / age / sex. If the coefficient shrinks substantially, those covariates were carrying part of the apparent disease signal.

```{r adjusted-models-species}
alpha_species_complete <- alpha_species |>
  filter(!is.na(BMI), !is.na(Host.age), !is.na(Sex))

mod_unadjusted_sp <- lm(Shannon ~ Disease.MESH.ID,
                        data = alpha_species_complete)
mod_adjusted_sp   <- lm(Shannon ~ Disease.MESH.ID + BMI + Host.age + Sex,
                        data = alpha_species_complete)

summary(mod_unadjusted_sp)$coefficients
summary(mod_adjusted_sp)$coefficients
```

# 9. Summary table

```{r summary-table}
summary_tbl <- wilcox_results |>
  left_join(effect_results, by = c("level", "metric"))

summary_tbl
```

# 10. Interpretation

Reading this table:

- For each metric × level, the Wilcoxon `p_value` tells you whether T2D and Healthy distributions differ.
- The `r` (rank-biserial) tells you how large the difference is. Positive r typically means the first level (T2D, D003924) tends to higher values; sign convention varies, so always check by comparing medians.
- `median_T2D` and `median_HC` give the direction of the difference at a glance.
- Whether the same direction and significance hold at both genus and species levels is itself informative — disagreement between levels often reflects a finer-resolution structure (e.g. genus-level signal that is not driven by a particular dominant species).

The covariate-adjusted linear model tells you whether the Disease coefficient survives BMI / age / sex adjustment. A surviving coefficient is the cleanest claim of a disease-specific alpha-diversity effect.

# 11. Save outputs

```{r save-outputs}
#saveRDS(alpha_all,        "project/results/alpha_diversity_table.rds")
#saveRDS(summary_tbl,      "project/results/alpha_diversity_summary.rds")
#saveRDS(mod_adjusted,     "project/results/alpha_lm_adjusted_genus.rds")
#saveRDS(mod_adjusted_sp,  "project/results/alpha_lm_adjusted_species.rds")
```

(If `project/results/` does not exist, create it once with `dir.create("project/results")`.)
