---
title: "Alpha diversity — filtered cohort"
author: "Anton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Alpha diversity within the filtered Chinese cohort (`Project.ID == "PRJNA422434"`, 187 T2D + 172 Healthy). All analyses run at both genus and species level. The headline test is Wilcoxon rank-sum; group medians, depth confounding, and covariate adjustment are also reported.

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
```

> *Taught in: W20-02 (`02_statistical_testing_of_alpha_diversity.Rmd`) §1 — phyloseq + tidyverse for alpha-diversity testing; ggplot2 from the W20 diversity practicals.*

# 2. Load and filter data

Load the original phyloseq objects and restrict to the Chinese cohort (`Project.ID == "PRJNA422434"`). The rationale for this filter is documented separately; in short, the other two projects in the dataset are case-only and cannot contribute to a case-control comparison.

```{r load-filter}
# Genus level
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps_genus_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

# Species level
load("project/data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData")
ps_species_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps_genus_clean
ps_species_clean
```

> *Taught in: W20-01/W20-02 — loading the course phyloseq objects; `subset_samples()` is the phyloseq subsetting shown in W20-03 (`03_beta_diversity_and_ordination.Rmd`). Cohort filter (PRJNA422434 only) is project-chosen, rationale in `cohort_rationale.Rmd`.*

# 3. Build a tidy alpha-diversity table at both levels

```{r alpha-table}
make_alpha_df <- function(ps, level_label) {
  alpha <- estimate_richness(ps, measures = c("Shannon", "Simpson", "InvSimpson")) |>
    rownames_to_column("sample_id")

  meta <- sample_data(ps) |>
    data.frame() |>
    rownames_to_column("sample_id")

  inner_join(alpha, meta, by = "sample_id") |>
    mutate(level = level_label)
}

alpha_genus   <- make_alpha_df(ps_genus_clean,   "Genus")
alpha_species <- make_alpha_df(ps_species_clean, "Species")
alpha_all     <- bind_rows(alpha_genus, alpha_species)

head(alpha_all)
```

> *Taught in: W20-02 §2 — `estimate_richness()` for Shannon/Simpson/InvSimpson. Joining the richness table to `sample_data()` follows the same practical.*

Note: `Observed` is omitted because the OTU table contains relative abundances rather than integer counts, which is the data structure `Observed`/`Chao1`/`ACE` assume.

# 4. Wilcoxon tests per metric and level

```{r wilcoxon-tests}
metrics_to_test <- c("Shannon", "Simpson", "InvSimpson")
levels_to_test  <- c("Genus", "Species")

wilcox_results <- tibble()

for (lvl in levels_to_test) {
  sub <- alpha_all |> filter(level == lvl)
  for (m in metrics_to_test) {
    w <- wilcox.test(reformulate("Disease.MESH.ID", m), data = sub)
    wilcox_results <- bind_rows(
      wilcox_results,
      tibble(
        level     = lvl,
        metric    = m,
        W         = unname(w$statistic),
        p_value   = w$p.value,
        median_T2D = median(sub[[m]][sub$Disease.MESH.ID == "D003924"]),
        median_HC  = median(sub[[m]][sub$Disease.MESH.ID == "D006262"])
      )
    )
  }
}

wilcox_results <- wilcox_results |>
  mutate(p_value = signif(p_value, 3))

wilcox_results
```

> *Taught in: W20-02 — Wilcoxon rank-sum (`wilcox.test`) / Kruskal-Wallis for two-group alpha-diversity comparison; group medians are the descriptive summary shown alongside the test in the same practical.*

# 5. Group medians and spread (IQR)

The Wilcoxon test above is non-parametric, so the matching descriptive summary is the group median and interquartile range (IQR) rather than a parametric effect size. This describes the size and direction of the difference using the same rank-based logic as the test.

```{r median-iqr}
median_iqr <- tibble()

for (lvl in levels_to_test) {
  sub <- alpha_all |> filter(level == lvl)
  for (m in metrics_to_test) {
    t2d <- sub[[m]][sub$Disease.MESH.ID == "D003924"]
    hc  <- sub[[m]][sub$Disease.MESH.ID == "D006262"]
    median_iqr <- bind_rows(
      median_iqr,
      tibble(
        level      = lvl,
        metric     = m,
        median_T2D = median(t2d),
        IQR_T2D    = IQR(t2d),
        median_HC  = median(hc),
        IQR_HC     = IQR(hc),
        median_diff = median(t2d) - median(hc)
      )
    )
  }
}

median_iqr <- median_iqr |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))

median_iqr
```

> *Taught in: W20-02 — median/IQR are the standard descriptive summary paired with the Wilcoxon test in the alpha-diversity practical. (Rank-biserial effect size from the `effectsize` package was removed: not part of the course material.)*

# 6. Depth confounding

A diversity difference could in principle reflect differential sequencing depth between groups. Spearman correlation between depth (`Nr..reads.sequenced`) and Shannon diversity tests this directly.

```{r depth-check}
depth_results <- alpha_all |>
  group_by(level) |>
  summarise(
    rho      = cor(Nr..reads.sequenced, Shannon, method = "spearman"),
    p_value  = cor.test(Nr..reads.sequenced, Shannon, method = "spearman")$p.value
  ) |>
  mutate(across(c(rho, p_value), \(x) signif(x, 3)))

depth_results
```

> *Taught in: W20-02 — sequencing-depth confounding exercise: Spearman `cor.test(..., method = "spearman")` of depth against a diversity metric. (Here depth = `Nr..reads.sequenced` from sample metadata.)*

A small |ρ| (< 0.1) and a non-significant p-value indicate no meaningful depth confound.

# 7. Visualise alpha diversity by disease

```{r plot-alpha, fig.width=9, fig.height=6}
alpha_long <- alpha_all |>
  select(sample_id, level, Disease.MESH.ID,
         Shannon, Simpson, InvSimpson) |>
  pivot_longer(cols = c(Shannon, Simpson, InvSimpson),
               names_to = "metric", values_to = "value") |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

ggplot(alpha_long, aes(x = Disease, y = value, fill = Disease)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.8) +
  facet_grid(metric ~ level, scales = "free_y") +
  scale_fill_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#EE4B2B")) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Alpha diversity",
       title = "Alpha diversity in the Chinese cohort",
       subtitle = "Shannon, Simpson, Inverse Simpson at genus and species level")
```

> *Taught in: W20-02 — boxplots of alpha diversity by group (`plot_richness` / ggplot boxplots) in the alpha-diversity practical.*

# 8. Covariate adjustment

Within the filtered cohort, additional covariates may still confound the disease effect — BMI, host age, and sex are biologically plausible candidates. Wilcoxon does not natively adjust for covariates, so we switch to a linear model on Shannon diversity. This is reasonable because Shannon at sample level is approximately continuous and not extremely skewed.

```{r covariate-check}
# Inspect covariate availability
summary(alpha_genus$BMI)
summary(alpha_genus$Host.age)
table(alpha_genus$Sex, alpha_genus$Disease.MESH.ID, useNA = "ifany")
```

> *Taught in: W21 reference Rmd §7 / Alberdi reading ("Multivariate statistics and multiple correction") §5 — covariate inspection prior to adjustment. BMI / host age / sex are project-chosen covariates (the method of adjusting for covariates is taught; the specific choice of these three is ours).*

If `BMI` has many NAs, those samples are dropped from the adjusted model only — the unadjusted comparison still uses all samples.

```{r adjusted-models}
alpha_genus_complete <- alpha_genus |>
  filter(!is.na(BMI), !is.na(Host.age), !is.na(Sex))

mod_unadjusted <- lm(Shannon ~ Disease.MESH.ID,
                     data = alpha_genus_complete)

mod_adjusted   <- lm(Shannon ~ Disease.MESH.ID + BMI + Host.age + Sex,
                     data = alpha_genus_complete)

summary(mod_unadjusted)$coefficients
summary(mod_adjusted)$coefficients
```

> *Taught in: W21 reference Rmd §7 / Alberdi reading §5 — linear models with covariates to adjust a group effect. Applied here to Shannon diversity; BMI / host age / sex are the project-chosen covariates.*

The key comparison is the `Disease.MESH.IDD006262` coefficient between the unadjusted and adjusted models. If the magnitude is similar and the term remains significant after adjustment, the disease effect is not explained by BMI / age / sex. If the coefficient shrinks substantially, those covariates were carrying part of the apparent disease signal.

```{r adjusted-models-species}
alpha_species_complete <- alpha_species |>
  filter(!is.na(BMI), !is.na(Host.age), !is.na(Sex))

mod_unadjusted_sp <- lm(Shannon ~ Disease.MESH.ID,
                        data = alpha_species_complete)
mod_adjusted_sp   <- lm(Shannon ~ Disease.MESH.ID + BMI + Host.age + Sex,
                        data = alpha_species_complete)

summary(mod_unadjusted_sp)$coefficients
summary(mod_adjusted_sp)$coefficients
```

> *Taught in: W21 reference Rmd §7 / Alberdi reading §5 — same covariate-adjusted linear model repeated at species level (project direction: report both genus and species).*

# 9. Summary table

```{r summary-table}
summary_tbl <- wilcox_results |>
  left_join(median_iqr, by = c("level", "metric", "median_T2D", "median_HC"))

summary_tbl
```

> *Taught in: project synthesis — joins the Wilcoxon p-values (W20-02) with the median/IQR descriptive summary (W20-02).*

# 10. Interpretation

Reading this table:

- For each metric × level, the Wilcoxon `p_value` tells you whether T2D and Healthy distributions differ.
- `median_T2D`, `median_HC`, and `median_diff` give the size and direction of the difference at a glance; `IQR_T2D` / `IQR_HC` describe the spread within each group.
- Whether the same direction and significance hold at both genus and species levels is itself informative — disagreement between levels often reflects a finer-resolution structure (e.g. genus-level signal that is not driven by a particular dominant species). Per the teacher's guidance, a genus-level difference is not "less real" than a species-level one and is not disregarded when species is non-significant.

The covariate-adjusted linear model tells you whether the Disease coefficient survives BMI / age / sex adjustment. A surviving coefficient is the cleanest claim of a disease-specific alpha-diversity effect.

# 11. Save outputs

```{r save-outputs}
#saveRDS(alpha_all,        "project/results/alpha_diversity_table.rds")
#saveRDS(summary_tbl,      "project/results/alpha_diversity_summary.rds")
#saveRDS(mod_adjusted,     "project/results/alpha_lm_adjusted_genus.rds")
#saveRDS(mod_adjusted_sp,  "project/results/alpha_lm_adjusted_species.rds")
```

> *Taught in: housekeeping (`saveRDS`/`dir.create`) — not a course method; persists results for reuse.*

(If `project/results/` does not exist, create it once with `dir.create("project/results")`.)
