---
title: "Beta diversity — filtered cohort"
author: "Anton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Purpose

Beta diversity within the filtered Chinese cohort (`Project.ID == "PRJNA422434"`, 187 T2D + 172 Healthy). Three distance metrics are computed in parallel because each emphasises a different aspect of community structure:

| Metric | Input data | Aspect captured | Compositionally correct? |
|---|---|---|---|
| **Bray–Curtis** | Relative abundance | Abundance differences | No |
| **Jaccard** | Presence/absence | Membership (which taxa are shared) | N/A (binary) |
| **Aitchison** | CLR-transformed | Compositional log-ratio differences | **Yes** (Aitchison §7) |

For each metric we run PCoA, PERMANOVA, and `betadisper`. PERMANOVA tests whether group centroids differ; `betadisper` tests whether within-group dispersions are homogeneous. A clean centroid-shift interpretation requires significant PERMANOVA with non-significant betadisper.

# 1. Libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(ggplot2)
library(dplyr)
library(tibble)
```

# 2. Load and filter data

Load the original phyloseq objects and restrict to the Chinese cohort (`Project.ID == "PRJNA422434"`). The rationale for this filter is documented separately; in short, the other two projects in the dataset are case-only and cannot contribute to a case-control comparison.

```{r load-filter}
# Genus level
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps_genus_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

# Species level
load("project/data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData")
ps_species_clean <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps_genus_clean
ps_species_clean
```

# 3. Helper functions

```{r helpers}
# Prepare phyloseq transformations
prepare_ps <- function(ps) {
  list(
    rel = transform_sample_counts(ps, function(x) x / sum(x)),
    pa  = transform_sample_counts(ps, function(x) ifelse(x > 0, 1, 0))
  )
}

# Compute Aitchison distance via vegdist (handles pseudocount + CLR internally)
aitchison_distance <- function(ps) {
  otu <- as(otu_table(ps), "matrix")
  if (taxa_are_rows(ps)) otu <- t(otu)
  vegdist(otu, method = "aitchison", pseudocount = 1)
}
```

# 4. Compute all distance matrices

```{r compute-distances}
g <- prepare_ps(ps_genus_clean)
s <- prepare_ps(ps_species_clean)

# Genus level
bray_g    <- distance(g$rel, method = "bray")
jaccard_g <- distance(g$pa,  method = "jaccard", binary = TRUE)
aitch_g   <- aitchison_distance(ps_genus_clean)

# Species level
bray_s    <- distance(s$rel, method = "bray")
jaccard_s <- distance(s$pa,  method = "jaccard", binary = TRUE)
aitch_s   <- aitchison_distance(ps_species_clean)
```

# 5. PCoA — all three metrics, both levels

```{r pcoa-all, fig.width=10, fig.height=10}
make_ord_df <- function(ps_for_plot, dist_obj, metric_label, level_label) {
  ord <- ordinate(ps_for_plot, method = "PCoA", distance = dist_obj)
  eig <- ord$values$Relative_eig
  df  <- plot_ordination(ps_for_plot, ord, type = "samples", justDF = TRUE) |>
    mutate(metric = metric_label,
           level  = level_label,
           ax1_pct = round(100 * eig[1], 1),
           ax2_pct = round(100 * eig[2], 1))
  df
}

ord_dfs <- bind_rows(
  make_ord_df(g$rel, bray_g,    "Bray-Curtis", "Genus"),
  make_ord_df(g$pa,  jaccard_g, "Jaccard",     "Genus"),
  make_ord_df(ps_genus_clean, aitch_g, "Aitchison", "Genus"),
  make_ord_df(s$rel, bray_s,    "Bray-Curtis", "Species"),
  make_ord_df(s$pa,  jaccard_s, "Jaccard",     "Species"),
  make_ord_df(ps_species_clean, aitch_s, "Aitchison", "Species")
) |>
  mutate(Disease = ifelse(Disease.MESH.ID == "D003924", "T2D", "Healthy"))

ggplot(ord_dfs, aes(x = Axis.1, y = Axis.2, color = Disease)) +
  geom_point(size = 1.8, alpha = 0.7) +
  stat_ellipse(level = 0.95) +
  facet_grid(level ~ metric, scales = "free") +
  scale_color_manual(values = c("Healthy" = "#56B4E9", "T2D" = "#E69F00")) +
  theme_bw() +
  labs(title = "PCoA across three distance metrics, two taxonomic levels",
       subtitle = "Filtered cohort PRJNA422434 (n = 359)",
       x = "PCoA 1", y = "PCoA 2")
```

# 6. PERMANOVA — all six combinations

```{r permanova-all}
run_permanova <- function(dist_obj, sample_df, level_label, metric_label) {
  ad <- adonis2(dist_obj ~ Disease.MESH.ID, data = sample_df, permutations = 999)
  tibble(
    level   = level_label,
    metric  = metric_label,
    F_stat  = round(ad$F[1], 2),
    R2      = round(ad$R2[1], 4),
    p_value = ad$`Pr(>F)`[1]
  )
}

sample_df_g <- sample_data(ps_genus_clean)   |> data.frame()
sample_df_s <- sample_data(ps_species_clean) |> data.frame()

permanova_results <- bind_rows(
  run_permanova(bray_g,    sample_df_g, "Genus",   "Bray-Curtis"),
  run_permanova(jaccard_g, sample_df_g, "Genus",   "Jaccard"),
  run_permanova(aitch_g,   sample_df_g, "Genus",   "Aitchison"),
  run_permanova(bray_s,    sample_df_s, "Species", "Bray-Curtis"),
  run_permanova(jaccard_s, sample_df_s, "Species", "Jaccard"),
  run_permanova(aitch_s,   sample_df_s, "Species", "Aitchison")
)

permanova_results
```

# 7. betadisper — dispersion check for each PERMANOVA

```{r betadisper-all}
run_betadisper <- function(dist_obj, grouping, level_label, metric_label) {
  disp <- betadisper(dist_obj, grouping)
  aov_out <- anova(disp)
  tibble(
    level   = level_label,
    metric  = metric_label,
    F_stat  = round(aov_out$`F value`[1], 2),
    p_value = aov_out$`Pr(>F)`[1]
  )
}

betadisper_results <- bind_rows(
  run_betadisper(bray_g,    sample_df_g$Disease.MESH.ID, "Genus",   "Bray-Curtis"),
  run_betadisper(jaccard_g, sample_df_g$Disease.MESH.ID, "Genus",   "Jaccard"),
  run_betadisper(aitch_g,   sample_df_g$Disease.MESH.ID, "Genus",   "Aitchison"),
  run_betadisper(bray_s,    sample_df_s$Disease.MESH.ID, "Species", "Bray-Curtis"),
  run_betadisper(jaccard_s, sample_df_s$Disease.MESH.ID, "Species", "Jaccard"),
  run_betadisper(aitch_s,   sample_df_s$Disease.MESH.ID, "Species", "Aitchison")
)

betadisper_results
```

# 8. Joint summary table

```{r joint-table}
beta_summary <- permanova_results |>
  rename(perm_F = F_stat, perm_p = p_value) |>
  left_join(
    betadisper_results |>
      rename(disp_F = F_stat, disp_p = p_value),
    by = c("level", "metric")
  ) |>
  mutate(
    perm_p_label = ifelse(perm_p < 0.001, "< 0.001", signif(perm_p, 3)),
    disp_p_label = ifelse(disp_p < 0.001, "< 0.001", signif(disp_p, 3)),
    cleanly_interpretable = perm_p < 0.05 & disp_p > 0.05
  )

beta_summary
```

The `cleanly_interpretable` column flags rows where a significant PERMANOVA result can be interpreted as a true centroid shift (significant PERMANOVA AND non-significant betadisper).

# 9. Covariate-adjusted PERMANOVA

Within the filtered cohort, residual confounders (BMI, age, sex) may still partly explain community variation. The marginal PERMANOVA partitions variance between each predictor adjusted for the others.

```{r adjusted-permanova}
adjusted_df_g <- sample_df_g |>
  filter(!is.na(BMI), !is.na(Host.age), !is.na(Sex))

# Re-subset distance matrix to match
keep_g <- sample_df_g$Run.ID %in% adjusted_df_g$Run.ID
bray_g_keep <- as.matrix(bray_g)[keep_g, keep_g] |> as.dist()

adonis2(bray_g_keep ~ Disease.MESH.ID + BMI + Host.age + Sex,
        data = adjusted_df_g,
        by = "margin",
        permutations = 999)
```

```{r adjusted-permanova-aitchison}
aitch_g_keep <- as.matrix(aitch_g)[keep_g, keep_g] |> as.dist()

adonis2(aitch_g_keep ~ Disease.MESH.ID + BMI + Host.age + Sex,
        data = adjusted_df_g,
        by = "margin",
        permutations = 999)
```

Aitchison adjusted is the most defensible single test — compositionally correct AND covariate adjusted. If `Disease.MESH.ID` is significant in this model, the disease effect on community composition survives adjustment for BMI, age, and sex within the cohort.

# 10. Interpretation guide

Reading the joint summary:

- **All three metrics significant in PERMANOVA** ⇒ robust disease effect across abundance-based, presence/absence, and compositional perspectives.
- **Aitchison `cleanly_interpretable = TRUE`** ⇒ the compositionally-rigorous metric gives a clean centroid-shift interpretation.
- **Disagreement between genus and species** ⇒ informative; the level at which the signal appears tells you about the resolution of the disease effect.
- **R² values** ⇒ typically 1–5% for disease in human gut microbiome studies; report alongside p-values to avoid over-claiming statistical significance for small effects.

For the report, the cleanest single sentence is: *"PERMANOVA on Aitchison distance, adjusted for BMI, host age, and sex, partitioned X% of community variance to disease state (p = Y) with no dispersion confound (betadisper p = Z)."*

# 11. Save outputs

```{r save-outputs}
#saveRDS(list(bray = bray_g, jaccard = jaccard_g, aitchison = aitch_g),
#        "project/results/beta_distances_genus.rds")

#saveRDS(list(bray = bray_s, jaccard = jaccard_s, aitchison = aitch_s),
#        "project/results/beta_distances_species.rds")

#saveRDS(beta_summary, "project/results/beta_diversity_summary.rds")
```
