---
title: "Differential abundance analysis — T2D vs Healthy (Chinese cohort)"
author: "Anton"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This Rmd mirrors the structure of the Week 21 practical (Antton Alberdi's reference Rmd), adapted to the T2D vs Healthy comparison within the filtered Chinese cohort (`Project.ID == "PRJNA422434"`).

| Teacher's variable | Used here |
|---|---|
| `body_site == "gut"` | (already filtered to Chinese cohort; no body-site variable) |
| `age` (adult, infant) | `Disease.MESH.ID` (D003924 = T2D, D006262 = Healthy) |
| `status` (healthy, sick) | `Sex` (Female, Male) — used as adjustment covariate |
| `genus`, `species` (lowercase) | `Genus`, `Species` (uppercase) — note: the `Species` column in this dataset actually contains genus names (column-shift quirk) |

Direction convention throughout: **positive** estimate = T2D-enriched, **negative** estimate = T2D-depleted (= Healthy-enriched).

## Loading of required libraries

```{r libraries}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggh4x)
library(broom)
library(microbiome)
```

## Load and inspect the data

```{r load-data}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
sample_variables(ps)
rank_names(ps)
table(sample_data(ps)$Disease.MESH.ID)
```

Expected: 187 T2D + 172 Healthy = 359 samples, all from PRJNA422434.

## Visualise community composition

```{r composition-bar, fig.width=11, fig.height=5}
ps_rel  <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- data.frame(sample_data(ps_rel))
tax_rel  <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()

plot_bar(ps_rel, fill = "Species") +
  geom_bar(aes(fill = Species), stat = "identity", colour = NA) +
  facet_nested(~ Sex + Disease.MESH.ID, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none") +
  labs(title = "Genus-level composition by Sex × Disease",
       fill = "Genus")
```

The data is already at genus level (58 taxa) — no `tax_glom` needed. Legend hidden because 58 categories are unreadable.

```{r build-dat-rel, include=FALSE}
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel  %>% rownames_to_column("taxon"),     by = "taxon")
```

# Univariate statistics

## Explore individual taxa

```{r single-taxon, fig.width=10, fig.height=4}
one_taxon <- unique(dat_rel$taxon)[3]
# one_taxon <- unique(dat_rel$taxon)[5]
# one_taxon <- unique(dat_rel$taxon)[20]

# Comparison between disease groups
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    theme_bw() +
    labs(title = paste("Taxon", one_taxon, "by disease"))

# Comparison between sexes within T2D
dat_rel %>%
  filter(taxon == one_taxon, Disease.MESH.ID == "D003924") %>%
  ggplot(aes(x = Sex, y = rel_abund, fill = Sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    theme_bw() +
    labs(title = paste("Taxon", one_taxon, "by sex (T2D only)"))
```

## Diagnose the data before testing

### Zero-inflation

```{r zero-inflation, fig.width=10, fig.height=4}
dat_rel %>%
  group_by(taxon, Disease.MESH.ID) %>%
  summarise(prop_zero = mean(rel_abund == 0), .groups = "drop") %>%
  ggplot(aes(x = prop_zero, fill = Disease.MESH.ID)) +
    geom_histogram(binwidth = 0.05, color = "white", alpha = 0.7) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    facet_wrap(~ Disease.MESH.ID) +
    labs(x = "Proportion of samples where relative abundance = 0",
         y = "Number of taxa",
         title = "Zero-inflation by disease group") +
    theme_bw() +
    theme(legend.position = "none")
```

### Right skew on a sample of mid-prevalence taxa

```{r right-skew, fig.width=10, fig.height=6}
set.seed(42)
example_taxa <- dat_rel %>%
  filter(Disease.MESH.ID == "D003924") %>%
  group_by(taxon, Genus, Species) %>%
  summarise(prop_nonzero = mean(rel_abund > 0), .groups = "drop") %>%
  filter(prop_nonzero > 0.4, prop_nonzero < 0.9) %>%
  slice_sample(n = 6) %>%
  pull(taxon)

dat_rel %>%
  filter(Disease.MESH.ID == "D003924",
         taxon %in% example_taxa,
         rel_abund > 0) %>%
  mutate(label = Species) %>%
  ggplot(aes(x = rel_abund)) +
    geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Relative abundance (non-zero values only)",
         y = "Count",
         title = "Right-skewed distributions in selected genera (T2D)") +
    theme_bw()
```

### Q-Q plots

```{r qq-plots, fig.width=10, fig.height=6}
dat_rel %>%
  filter(Disease.MESH.ID == "D003924", taxon %in% example_taxa) %>%
  mutate(label = Species) %>%
  ggplot(aes(sample = rel_abund)) +
    stat_qq(size = 0.8, alpha = 0.6) +
    stat_qq_line(color = "red", linetype = "dashed") +
    facet_wrap(~ label, scales = "free") +
    labs(x = "Theoretical quantiles (normal)",
         y = "Sample quantiles",
         title = "Q–Q plots: departures from normality in selected genera (T2D)") +
    theme_bw()
```

Zero-inflation, right skew and Q-Q curvature together justify rank-based or compositional methods over raw t-tests.

## Wilcoxon rank-sum test

```{r wilcox-results}
wilcox_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    p_value      = wilcox.test(rel_abund ~ Disease.MESH.ID, exact = FALSE)$p.value,
    mean_T2D     = mean(rel_abund[Disease.MESH.ID == "D003924"]),
    mean_Healthy = mean(rel_abund[Disease.MESH.ID == "D006262"]),
    log2_fc      = log2((mean_T2D + 1e-6) / (mean_Healthy + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

head(wilcox_results, 10)
```

`log2_fc > 0` → T2D-enriched. `log2_fc < 0` → T2D-depleted.

### Top T2D-depleted (Wilcoxon)

```{r wilcox-depleted}
wilcox_T2D_depleted <- wilcox_results %>%
  select(taxon, Genus, Species, mean_T2D, mean_Healthy, log2_fc, p_value, q_value) %>%
  arrange(log2_fc) %>%
  head(10)

wilcox_T2D_depleted
```

```{r wilcox-depleted-plot, fig.width=6, fig.height=4}
top_taxon <- wilcox_T2D_depleted %>% slice(1) %>% pull(taxon)
top_label <- wilcox_T2D_depleted %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-depleted (Wilcoxon):", top_label))
```

### Top T2D-enriched (Wilcoxon)

```{r wilcox-enriched}
wilcox_T2D_enriched <- wilcox_results %>%
  select(taxon, Genus, Species, mean_T2D, mean_Healthy, log2_fc, p_value, q_value) %>%
  arrange(-log2_fc) %>%
  head(10)

wilcox_T2D_enriched
```

```{r wilcox-enriched-plot, fig.width=6, fig.height=4}
top_taxon <- wilcox_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- wilcox_T2D_enriched %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-enriched (Wilcoxon):", top_label))
```

### Volcano plot — Wilcoxon

```{r wilcox-volcano, fig.width=8, fig.height=5}
wilcox_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc > 0 ~ "T2D-enriched",
      q_value < 0.05 & log2_fc < 0 ~ "T2D-depleted",
      TRUE                         ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "log2 fold change (T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "Wilcoxon: disease effect on relative abundance") +
    theme_bw()
```

### Bar chart of top Wilcoxon hits

```{r wilcox-bar, fig.width=8, fig.height=6}
bind_rows(
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(log2_fc)  %>% head(10),
  wilcox_results %>% filter(q_value < 0.05) %>% arrange(-log2_fc) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = Species,
    direction = ifelse(log2_fc > 0, "T2D-enriched", "T2D-depleted")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = log2_fc, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "log2 fold change (T2D vs Healthy)", fill = NULL,
         title = "Top differentially abundant taxa (Wilcoxon)") +
    theme_bw()
```

## Disease effect using log-transformed linear models

```{r glm-results}
glm_disease_results <- dat_rel %>%
  mutate(Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924"))) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID,
               data = .x, family = gaussian())
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_T2D_vs_Healthy = 10^estimate
  )
```

Positive estimate = T2D-enriched on the log10 scale.

### Top T2D-depleted (GLM)

```{r glm-depleted}
glm_T2D_depleted <- glm_disease_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy,
         std.error, statistic, p.value, q_value) %>%
  arrange(fold_change_T2D_vs_Healthy) %>%
  head(10)

glm_T2D_depleted
```

```{r glm-depleted-plot, fig.width=6, fig.height=4}
top_taxon <- glm_T2D_depleted %>% slice(1) %>% pull(taxon)
top_label <- glm_T2D_depleted %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-depleted (GLM):", top_label))
```

### Top T2D-enriched (GLM)

```{r glm-enriched}
glm_T2D_enriched <- glm_disease_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy,
         std.error, statistic, p.value, q_value) %>%
  arrange(-fold_change_T2D_vs_Healthy) %>%
  head(10)

glm_T2D_enriched
```

```{r glm-enriched-plot, fig.width=6, fig.height=4}
top_taxon <- glm_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- glm_T2D_enriched %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-enriched (GLM):", top_label))
```

### Volcano plot — GLM (unadjusted)

```{r glm-volcano, fig.width=8, fig.height=5}
glm_disease_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
      TRUE                          ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "GLM: disease effect on log10 relative abundance") +
    theme_bw()
```

### Bar chart — GLM top hits

```{r glm-bar, fig.width=8, fig.height=6}
bind_rows(
  glm_disease_results %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(10),
  glm_disease_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = Species,
    direction = ifelse(estimate > 0, "T2D-enriched", "T2D-depleted")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "GLM estimate (log10 RA)", fill = NULL,
         title = "Top differentially abundant taxa (GLM)") +
    theme_bw()
```

## Adjust the linear model for Sex

```{r glm-adj-results}
glm_disease_adjusted_sex_results <- dat_rel %>%
  mutate(
    Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
    Sex             = factor(Sex,             levels = c("Female", "Male"))
  ) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID + Sex,
               data = .x, family = gaussian())
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_T2D_vs_Healthy_adj = 10^estimate
  )
```

### Top T2D-depleted (GLM adjusted for Sex)

```{r glm-adj-depleted}
glm_adj_T2D_depleted <- glm_disease_adjusted_sex_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy_adj,
         std.error, statistic, p.value, q_value) %>%
  arrange(fold_change_T2D_vs_Healthy_adj) %>%
  head(10)

glm_adj_T2D_depleted
```

```{r glm-adj-depleted-plot, fig.width=6, fig.height=4}
top_taxon <- glm_adj_T2D_depleted %>% slice(1) %>% pull(taxon)
top_label <- glm_adj_T2D_depleted %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-depleted (GLM adj. Sex):", top_label))
```

### Top T2D-enriched (GLM adjusted for Sex)

```{r glm-adj-enriched}
glm_adj_T2D_enriched <- glm_disease_adjusted_sex_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy_adj,
         std.error, statistic, p.value, q_value) %>%
  arrange(-fold_change_T2D_vs_Healthy_adj) %>%
  head(10)

glm_adj_T2D_enriched
```

```{r glm-adj-enriched-plot, fig.width=6, fig.height=4}
top_taxon <- glm_adj_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- glm_adj_T2D_enriched %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-enriched (GLM adj. Sex):", top_label))
```

### Volcano plot — GLM adjusted for Sex

```{r glm-adj-volcano, fig.width=8, fig.height=5}
glm_disease_adjusted_sex_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
      TRUE                          ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "GLM adjusted for Sex: disease effect") +
    theme_bw()
```

### Bar chart — GLM adjusted top hits

```{r glm-adj-bar, fig.width=8, fig.height=6}
bind_rows(
  glm_disease_adjusted_sex_results %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(10),
  glm_disease_adjusted_sex_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = Species,
    direction = ifelse(estimate > 0, "T2D-enriched", "T2D-depleted")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "GLM estimate (log10 RA, adjusted)", fill = NULL,
         title = "Top differentially abundant taxa (GLM adjusted for Sex)") +
    theme_bw()
```

## Compositional analysis with CLR transformation

```{r clr-transform}
ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})
otu_clr  <- as(otu_table(ps_clr), "matrix") %>% t() %>% as.data.frame()
meta_clr <- data.frame(sample_data(ps_clr))
tax_clr  <- as(tax_table(ps_clr), "matrix") %>% as.data.frame()
```

```{r build-dat-clr, include=FALSE}
dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "clr") %>%
  left_join(meta_clr %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_clr  %>% rownames_to_column("taxon"),     by = "taxon")
```

```{r clr-results}
clr_disease_adjusted_sex_results <- dat_clr %>%
  mutate(
    Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
    Sex             = factor(Sex,             levels = c("Female", "Male"))
  ) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- lm(clr ~ Disease.MESH.ID + Sex, data = .x)
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(q_value = p.adjust(p.value, method = "BH"))
```

Positive CLR estimate = T2D-enriched relative to the community geometric mean.

### Top T2D-depleted (CLR)

```{r clr-depleted}
clr_T2D_depleted <- clr_disease_adjusted_sex_results %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(estimate) %>%
  head(10)

clr_T2D_depleted
```

```{r clr-depleted-plot, fig.width=6, fig.height=4}
top_taxon <- clr_T2D_depleted %>% slice(1) %>% pull(taxon)
top_label <- clr_T2D_depleted %>% slice(1) %>% pull(Species)

dat_clr %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = clr, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-depleted (CLR):", top_label),
         y = "CLR-transformed value")
```

### Top T2D-enriched (CLR)

```{r clr-enriched}
clr_T2D_enriched <- clr_disease_adjusted_sex_results %>%
  select(taxon, Genus, Species, estimate, std.error, statistic, p.value, q_value) %>%
  arrange(-estimate) %>%
  head(10)

clr_T2D_enriched
```

```{r clr-enriched-plot, fig.width=6, fig.height=4}
top_taxon <- clr_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- clr_T2D_enriched %>% slice(1) %>% pull(Species)

dat_clr %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = clr, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-enriched (CLR):", top_label),
         y = "CLR-transformed value")
```

### Volcano plot — CLR

```{r clr-volcano, fig.width=8, fig.height=5}
clr_disease_adjusted_sex_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
      TRUE                          ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "CLR estimate (T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "Compositional (CLR) analysis: disease effect") +
    theme_bw()
```

### Bar chart — CLR top hits

```{r clr-bar, fig.width=8, fig.height=6}
bind_rows(
  clr_disease_adjusted_sex_results %>% filter(q_value < 0.05) %>% arrange(estimate)  %>% head(10),
  clr_disease_adjusted_sex_results %>% filter(q_value < 0.05) %>% arrange(-estimate) %>% head(10)
) %>%
  distinct() %>%
  mutate(
    label = Species,
    direction = ifelse(estimate > 0, "T2D-enriched", "T2D-depleted")
  ) %>%
  ggplot(aes(x = reorder(label, -q_value), y = estimate, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("T2D-enriched" = "#E74C3C", "T2D-depleted" = "#3498DB")) +
    labs(x = NULL, y = "CLR estimate (T2D vs Healthy)", fill = NULL,
         title = "Top differentially abundant taxa (CLR)") +
    theme_bw()
```

## Compare differential abundance methods

```{r comparison-build, include=FALSE}
comparison_results <- wilcox_results %>%
  select(taxon, Genus, Species,
         wilcox_log2fc = log2_fc,
         wilcox_q = q_value) %>%
  left_join(
    glm_disease_results %>%
      select(taxon, glm_estimate = estimate, glm_q = q_value),
    by = "taxon"
  ) %>%
  left_join(
    glm_disease_adjusted_sex_results %>%
      select(taxon, glm_adj_estimate = estimate, glm_adj_q = q_value),
    by = "taxon"
  ) %>%
  left_join(
    clr_disease_adjusted_sex_results %>%
      select(taxon, clr_estimate = estimate, clr_q = q_value),
    by = "taxon"
  )
```

### Significant taxa per method

```{r sig-summary}
sig_summary <- tibble(
  method = c("Wilcoxon", "GLM", "GLM adjusted (Sex)", "CLR adjusted (Sex)"),
  n_significant = c(
    sum(wilcox_results$q_value < 0.05, na.rm = TRUE),
    sum(glm_disease_results$q_value < 0.05, na.rm = TRUE),
    sum(glm_disease_adjusted_sex_results$q_value < 0.05, na.rm = TRUE),
    sum(clr_disease_adjusted_sex_results$q_value < 0.05, na.rm = TRUE)
  )
)

sig_summary
```

### Taxa significant in all four methods

```{r all-methods-sig}
all_methods_sig <- comparison_results %>%
  filter(wilcox_q < 0.05, glm_q < 0.05, glm_adj_q < 0.05, clr_q < 0.05) %>%
  select(Genus, Species, wilcox_q, glm_q, glm_adj_q, clr_q) %>%
  arrange(wilcox_q)

all_methods_sig
```

### GLM unadjusted vs adjusted for Sex

```{r glm-vs-glm-adj, fig.width=6, fig.height=5}
comparison_results %>%
  ggplot(aes(x = glm_estimate, y = glm_adj_estimate)) +
    geom_point(alpha = 0.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
    labs(x = "GLM estimate (unadjusted)",
         y = "GLM estimate (adjusted for Sex)",
         title = "Effect of adjusting for Sex") +
    theme_bw()
```

### GLM adjusted vs CLR

```{r glm-adj-vs-clr, fig.width=6, fig.height=5}
comparison_results %>%
  ggplot(aes(x = glm_adj_estimate, y = clr_estimate)) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "GLM estimate (adjusted, log10 RA)",
         y = "CLR estimate (adjusted)",
         title = "GLM vs compositional (CLR) effect sizes") +
    theme_bw()
```

### Concordance across methods

```{r concordance, fig.width=6, fig.height=4}
comparison_results %>%
  mutate(
    sig_wilcox  = wilcox_q  < 0.05,
    sig_glm     = glm_q     < 0.05,
    sig_glm_adj = glm_adj_q < 0.05,
    sig_clr     = clr_q     < 0.05,
    n_methods_significant = sig_wilcox + sig_glm + sig_glm_adj + sig_clr
  ) %>%
  count(n_methods_significant) %>%
  filter(!is.na(n_methods_significant)) %>%
  ggplot(aes(x = factor(n_methods_significant), y = n)) +
    geom_col(fill = "steelblue") +
    labs(x = "Number of methods detecting significance (q < 0.05)",
         y = "Number of taxa",
         title = "Concordance of significance across methods") +
    theme_bw()
```

### Significance × direction heatmap

```{r heatmap, fig.width=8, fig.height=8}
heatmap_data <- comparison_results %>%
  filter(wilcox_q < 0.05 | glm_q < 0.05 | glm_adj_q < 0.05 | clr_q < 0.05) %>%
  mutate(label = Species) %>%
  select(label,
         wilcox_q, glm_q, glm_adj_q, clr_q,
         wilcox_log2fc, glm_estimate, glm_adj_estimate, clr_estimate) %>%
  pivot_longer(cols = c(wilcox_q, glm_q, glm_adj_q, clr_q),
               names_to = "method", values_to = "q_value") %>%
  mutate(
    direction = case_when(
      method == "wilcox_q"  ~ wilcox_log2fc,
      method == "glm_q"     ~ glm_estimate,
      method == "glm_adj_q" ~ glm_adj_estimate,
      method == "clr_q"     ~ clr_estimate
    ),
    signed_neg_log10_q = sign(direction) * pmin(-log10(q_value), 10),
    method = factor(method,
                    levels = c("wilcox_q", "glm_q", "glm_adj_q", "clr_q"),
                    labels = c("Wilcoxon", "GLM", "GLM adj.", "CLR adj."))
  )

taxon_order <- heatmap_data %>%
  group_by(label) %>%
  summarise(mean_signed = mean(signed_neg_log10_q, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_signed) %>%
  pull(label)

heatmap_data %>%
  mutate(label = factor(label, levels = taxon_order)) %>%
  ggplot(aes(x = method, y = label, fill = signed_neg_log10_q)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = ifelse(q_value < 0.05, "*", "")),
              color = "black", size = 3) +
    scale_fill_gradient2(
      low = "#3498DB", mid = "white", high = "#C0392B",
      midpoint = 0, limits = c(-10, 10),
      name = "Signed\n-log10(q)",
      breaks = c(-10, -5, 0, 5, 10),
      labels = c("-10\n(T2D-depleted)", "-5", "0", "5", "10\n(T2D-enriched)")
    ) +
    labs(x = "Method", y = "Taxon",
         title = "Significance and direction across methods (* = q < 0.05)",
         subtitle = "Blue = T2D-depleted, Red = T2D-enriched") +
    theme_bw() +
    theme(axis.text.y = element_text(size = 7))
```

### Wilcoxon vs CLR — sign agreement

```{r wilcox-vs-clr, fig.width=7, fig.height=5}
comparison_results %>%
  mutate(
    sig_category = case_when(
      wilcox_q < 0.05 & clr_q < 0.05 ~ "Both significant",
      wilcox_q < 0.05                ~ "Wilcoxon only",
      clr_q < 0.05                   ~ "CLR only",
      TRUE                           ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = wilcox_log2fc, y = clr_estimate, color = sig_category)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c(
      "Both significant" = "#E74C3C",
      "Wilcoxon only"    = "#E67E22",
      "CLR only"         = "#3498DB",
      "Not significant"  = "grey70"
    )) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "Wilcoxon log2 FC (T2D vs Healthy)",
         y = "CLR estimate (T2D vs Healthy)",
         color = NULL,
         title = "Effect size agreement: Wilcoxon vs CLR") +
    theme_bw()
```

# Multivariate statistics

## PCA of CLR-transformed profiles

```{r pca, fig.width=7, fig.height=5}
clr_mat  <- otu_clr %>% as.matrix()
meta_pca <- meta_clr[rownames(clr_mat), ]

clr_pca <- prcomp(clr_mat, center = TRUE, scale. = FALSE)

pca_scores <- as.data.frame(clr_pca$x[, 1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_pca %>% rownames_to_column("sample_id"), by = "sample_id")

pca_var <- round(100 * clr_pca$sdev^2 / sum(clr_pca$sdev^2), 1)

pca_scores %>%
  ggplot(aes(x = PC1, y = PC2, color = Disease.MESH.ID, shape = Sex)) +
    geom_point(size = 3, alpha = 0.85) +
    scale_color_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    scale_shape_manual(values = c("Female" = 16, "Male" = 17)) +
    labs(x = paste0("PC1 (", pca_var[1], "%)"),
         y = paste0("PC2 (", pca_var[2], "%)"),
         color = "Disease", shape = "Sex",
         title = "PCA of CLR-transformed genus profiles") +
    theme_bw()
```

## Constrained ordination with RDA

```{r rda-fit}
rda_disease <- rda(clr_mat ~ Disease.MESH.ID + Sex, data = meta_pca)
```

### Permutation tests

```{r rda-anova}
anova(rda_disease, permutations = 999)
anova(rda_disease, by = "term", permutations = 999)
```

A significant disease term supports a community-level disease effect after accounting for sex.

### Top RDA1 taxa

```{r rda-taxa}
rda_site_scores_raw <- scores(rda_disease, display = "sites")
rda1_sign <- ifelse(
  mean(rda_site_scores_raw[meta_pca$Disease.MESH.ID == "D003924", "RDA1"], na.rm = TRUE) >=
    mean(rda_site_scores_raw[meta_pca$Disease.MESH.ID == "D006262", "RDA1"], na.rm = TRUE),
  1, -1
)

taxon_scores <- scores(rda_disease, display = "species")[, 1] * rda1_sign

rda_taxa <- tibble(
  taxon        = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```

### RDA biplot

```{r rda-biplot, fig.width=8, fig.height=6}
eig <- rda_disease$CCA$eig
pct1 <- round(eig[1] / rda_disease$tot.chi * 100, 1)
pct2 <- round(eig[2] / rda_disease$tot.chi * 100, 1)

site_scores <- scores(rda_disease, display = "sites") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("sample_id") %>%
  left_join(meta_pca %>% rownames_to_column("sample_id"), by = "sample_id")

biplot_scores <- scores(rda_disease, display = "bp") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("variable")

top_species_scores <- scores(rda_disease, display = "species") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("taxon") %>%
  inner_join(rda_taxa %>% head(10), by = "taxon") %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(label = Species)

site_range    <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))
arrow_scale   <- site_range * 0.8 / max(sqrt(biplot_scores$RDA1^2 + biplot_scores$RDA2^2))
species_scale <- site_range * 0.6 / max(sqrt(top_species_scores$RDA1^2 + top_species_scores$RDA2^2))

biplot_scores      <- biplot_scores      %>% mutate(RDA1 = RDA1 * arrow_scale,   RDA2 = RDA2 * arrow_scale)
top_species_scores <- top_species_scores %>% mutate(RDA1 = RDA1 * species_scale, RDA2 = RDA2 * species_scale)

ggplot() +
  geom_point(data = site_scores,
             aes(x = RDA1, y = RDA2, color = Disease.MESH.ID, shape = Sex),
             size = 3, alpha = 0.85) +
  geom_segment(data = top_species_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey70", linewidth = 0.5) +
  geom_text(data = top_species_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
            size = 2.5, color = "grey50") +
  geom_segment(data = biplot_scores,
               aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "black", linewidth = 1) +
  geom_text(data = biplot_scores,
            aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
            size = 3.5, fontface = "bold", color = "black") +
  scale_color_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
  scale_shape_manual(values = c("Female" = 16, "Male" = 17)) +
  labs(x = paste0("RDA1 (", pct1, "% of total variance)"),
       y = paste0("RDA2 (", pct2, "% of total variance)"),
       color = "Disease", shape = "Sex",
       title = "RDA biplot: gut microbiome constrained by Disease and Sex") +
  theme_bw()
```

### Top 20 taxa by RDA1 loading

```{r rda-loadings, fig.width=8, fig.height=6}
rda_taxa %>%
  head(20) %>%
  left_join(tax_clr %>% rownames_to_column("taxon"), by = "taxon") %>%
  mutate(
    label     = Species,
    direction = ifelse(rda1_loading > 0, "T2D-associated", "Healthy-associated")
  ) %>%
  ggplot(aes(x = reorder(label, rda1_loading), y = rda1_loading, fill = direction)) +
    geom_col() +
    coord_flip() +
    scale_fill_manual(values = c("T2D-associated"     = "#E74C3C",
                                 "Healthy-associated" = "#3498DB")) +
    labs(x = NULL, y = "RDA1 loading", fill = NULL,
         title = "Top 20 taxa by RDA1 loading magnitude") +
    theme_bw()
```

### RDA loadings vs CLR univariate estimates

```{r rda-vs-clr, fig.width=7, fig.height=5}
rda_clr_comparison <- rda_taxa %>%
  left_join(
    clr_disease_adjusted_sex_results %>%
      select(taxon, Genus, Species,
             clr_estimate = estimate,
             clr_q = q_value),
    by = "taxon"
  ) %>%
  mutate(
    direction_agreement = sign(rda1_loading) == sign(clr_estimate),
    clr_result = case_when(
      clr_q < 0.05 & direction_agreement ~ "CLR significant, same direction",
      clr_q < 0.05                       ~ "CLR significant, opposite direction",
      TRUE                               ~ "CLR not significant"
    )
  )

rda_clr_comparison %>%
  ggplot(aes(x = clr_estimate, y = rda1_loading, color = clr_result)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    geom_point(alpha = 0.65, size = 2) +
    scale_color_manual(values = c(
      "CLR significant, same direction"     = "#2E8B57",
      "CLR significant, opposite direction" = "#E67E22",
      "CLR not significant"                 = "grey70"
    )) +
    labs(x = "CLR disease estimate (T2D vs Healthy)",
         y = "RDA1 loading (positive = T2D-associated)",
         color = NULL,
         title = "RDA loadings vs CLR univariate disease effects") +
    theme_bw()
```

### Top 20 RDA loading taxa with their CLR statistics

```{r rda-clr-table}
rda_clr_comparison %>%
  arrange(desc(abs(rda1_loading))) %>%
  select(Genus, Species, rda1_loading, clr_estimate, clr_q, direction_agreement) %>%
  head(20)
```
