---
title: "Antons legestue genus niveau"
author: "Anton"
date: "2026-05-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Loading libraries
```{r}
library(phyloseq)
library(ggplot2)
library(vegan)
library(effectsize)
```

## loading dataset on genus level
```{r}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps_genus <- pylsq
# ps_genus <- subset_samples(ps_genus, Project.ID == "PRJNA422434") only china
rm(pylsq)
```

## Exploring
```{r}
ps_genus
sample_variables(ps_genus)
head(sample_data(ps_genus))


# how many are healthy/sick?
table(sample_data(ps_genus)$Disease.MESH.ID, useNA = "ifany")
```
## Shannon Rich?
```{r}
# Calculate Shannon diversity for all samples
shannon <- estimate_richness(
  ps_genus,
  measures = c("Shannon")
)

head(shannon)
```
## Visualizing shannon

```{r}
# Basic alpha-diversity plot
plot_richness(
  ps_genus,
  measures = c("Shannon")
)
```
## Grouping by sick/healthy
```{r, echo=FALSE}
# Group samples by health status
plot_richness(
  ps_genus,
  measures = "Shannon",
  x = "Disease.MESH.ID"
) +
  geom_boxplot(alpha = 0.6) +
  theme_bw()
```
### Sjovt med lavere diversitet hos raske. Måske det er fordi, det stadig er på genus level. Flere forskellige genus, men måske færre species pr. genus? HUSK AT TJEK!:)


## testing
```{r}
# Build the joined table
shannon_df <- shannon |>
  tibble::rownames_to_column("sample_id")

meta_df <- sample_data(ps_genus) |>
  data.frame() |>
  tibble::rownames_to_column("sample_id")

alpha_div <- dplyr::inner_join(shannon_df, meta_df, by = "sample_id")

# Test
wilcox.test(Shannon ~ Disease.MESH.ID, data = alpha_div)
```
## Now three methods in one
```{r}
alpha_metrics <- estimate_richness(
  ps_genus,
  measures = c("Shannon", "Simpson", "InvSimpson")
)
head(alpha_metrics)
```
## test for each

```{r}
# Reuse your alpha_div table, just rebuild with the new columns
alpha_metrics_df <- alpha_metrics |> tibble::rownames_to_column("sample_id")
alpha_div <- dplyr::inner_join(alpha_metrics_df, meta_df, by = "sample_id")

wilcox.test(Shannon    ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(Simpson    ~ Disease.MESH.ID, data = alpha_div)
wilcox.test(InvSimpson ~ Disease.MESH.ID, data = alpha_div)
```

## plotting

```{r}
plot_richness(
  ps_genus,
  measures = c("Shannon", "Simpson", "InvSimpson"),
  x = "Disease.MESH.ID"
) +
  geom_boxplot(alpha = 0.6) +
  theme_bw()
```

## T2D vs Healthy could differ in Shannon just because of different sequencing depths. Testing if thats the case
```{r}
# Spearman correlation between depth and Shannon
cor.test(alpha_div$Nr..reads.sequenced, alpha_div$Shannon, method = "spearman")

ggplot(alpha_div, aes(x = Nr..reads.sequenced, y = Shannon, color = Disease.MESH.ID)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  theme_bw()
```
Interpretation: All three diversity metrics — capturing different parts of the rare-to-dominant spectrum — agree. T2D patients show higher genus-level diversity than healthy controls, and this is NOT explained by differences in sequencing depth.

```{r}
rank_biserial(Shannon ~ Disease.MESH.ID, data = alpha_div)
```
# Transform to relative abundance
```{r}
ps_rel <- transform_sample_counts(ps_genus, function(x) x / sum(x))
# (your data is already relative abundance, but this normalizes any per-sample drift)
```
# Distance matrix
```{r}
bray_dist <- distance(ps_rel, method = "bray")
```
# Ordinate (PCoA)
```{r}
ord_bray <- ordinate(ps_rel, method = "PCoA", distance = bray_dist)
```

# Plot
```{r}
plot_ordination(ps_rel, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2) +
  stat_ellipse() +     # adds a 95% ellipse per group
  theme_bw()
```
```{r}
# Extract sample metadata
sample_df <- sample_data(ps_rel) |> data.frame()

# PERMANOVA — does composition differ between T2D and Healthy?
adonis_bray <- adonis2(
  bray_dist ~ Disease.MESH.ID,
  data = sample_df,
  permutations = 999
)
adonis_bray

# betadisper — are within-group dispersions equal?
dispersion_bray <- betadisper(bray_dist, sample_df$Disease.MESH.ID)
anova(dispersion_bray)
plot(dispersion_bray)
```

# 1. Check Project.ID as a confounder (still important from before):
```{r}
table(sample_data(ps_rel)$Country, sample_data(ps_rel)$Disease.MESH.ID)

table(sample_data(ps_rel)$Project.ID, sample_data(ps_rel)$Disease.MESH.ID)
```


```{r}
# Stratified PERMANOVA — test disease effect WITHIN each project
adonis2(
  bray_dist ~ Disease.MESH.ID,
  data = sample_df,
  strata = sample_df$Project.ID,    # restrict permutations within projects
  permutations = 999
)
```




