---
title: "Microbiome Analysis Workflow"
output: html_document
---

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Load libraries

```{r}
library(phyloseq)
library(ggplot2)
library(vegan)
library(dplyr)
library(tibble)
library(pheatmap)
library(ggh4x)
library(ggrepel)
library(microbiome)
```

# Load data

```{r}
load("~/group-project-group-6/project/data/D003924_Diabetes_Mellitus,_T/species/phyloseq.RData")

ps <- pylsq
rm(pylsq)
```

# Inspect phyloseq object

```{r}
ps
```

# Sample variables

```{r}
sample_variables(ps)
```

# Taxonomic ranks

```{r}
rank_names(ps)
```

# Disease distribution

```{r}
table(sample_data(ps)$Disease.MESH.ID)
```

# Project distribution

```{r}
table(sample_data(ps)$Project.ID)
```

# Disease by project

```{r}
table(sample_data(ps)$Project.ID,
      sample_data(ps)$Disease.MESH.ID)
```

# Country by project

```{r}
table(sample_data(ps)$Country,
      sample_data(ps)$Project.ID)
```

# Sample sums

```{r}
summary(sample_sums(ps))
```

# Histogram of sample sums

```{r}
hist(
  sample_sums(ps),
  breaks = 30,
  main = "Sample sums",
  xlab = "Sample sum"
)
```

# Relative abundance transformation

Although the data already appears normalized, this step rescales each sample to sum to 1.

```{r}
ps_rel <- transform_sample_counts(
  ps,
  function(x) x / sum(x)
)
```

# Bray-Curtis distance

```{r}
bray_dist <- distance(
  ps_rel,
  method = "bray"
)
```

# PCoA ordination

```{r}
ord_bray <- ordinate(
  ps_rel,
  method = "PCoA",
  distance = bray_dist
)
```

# PCoA colored by project

```{r fig.width=8, fig.height=5}
plot_ordination(
  ps_rel,
  ord_bray,
  color = "Project.ID"
) +
  geom_point(size = 3,
             alpha = 0.7) +
  stat_ellipse() +
  theme_bw() +
  labs(
    title = "PCoA by Project",
    subtitle = "Bray-Curtis distance"
  )
```

# PCoA colored by disease

```{r fig.width=8, fig.height=5}
plot_ordination(
  ps_rel,
  ord_bray,
  color = "Disease.MESH.ID",
  shape = "Project.ID"
) +
  geom_point(size = 3,
             alpha = 0.8) +
  stat_ellipse(aes(group = Disease.MESH.ID)) +
  theme_bw()
```

# PCoA faceted by project

```{r fig.width=10, fig.height=5}
plot_ordination(
  ps_rel,
  ord_bray,
  color = "Disease.MESH.ID"
) +
  geom_point(size = 3) +
  facet_wrap(~Project.ID) +
  theme_bw()
```

# PERMANOVA

```{r}
sample_df <- sample_data(ps_rel) |>
  data.frame()

adonis2(
  bray_dist ~ Project.ID + Disease.MESH.ID,
  data = sample_df,
  permutations = 999,
  by = "margin"
)
```

# Dispersion test

```{r}
disp_project <- betadisper(
  bray_dist,
  sample_df$Project.ID
)

anova(disp_project)

permutest(disp_project)
```

# Dispersion boxplot

```{r}
boxplot(disp_project)
```

# Alpha diversity

```{r}
alpha_div <- estimate_richness(
  ps_rel,
  measures = c("Shannon", "Simpson")
)

alpha_div$Disease <- sample_data(ps_rel)$Disease.MESH.ID
alpha_div$Project <- sample_data(ps_rel)$Project.ID
```

# Shannon diversity

```{r}
ggplot(
  alpha_div,
  aes(x = Disease,
      y = Shannon,
      fill = Disease)
) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2,
              alpha = 0.5) +
  theme_bw()
```

# Kruskal-Wallis test

```{r}
kruskal.test(
  Shannon ~ Disease,
  data = alpha_div
)
```

# Aggregate to genus level

```{r}
ps_genus <- tax_glom(
  ps_rel,
  taxrank = "Genus"
)
```

# Top 10 genera

```{r}
top10 <- names(sort(
  taxa_sums(ps_genus),
  decreasing = TRUE
))[1:10]

ps_top <- prune_taxa(
  top10,
  ps_genus
)
```

# Merge samples by disease

```{r}
ps_disease <- merge_samples(
  ps_top,
  group = "Disease.MESH.ID"
)

ps_disease <- transform_sample_counts(
  ps_disease,
  function(x) x / sum(x)
)
```

# Stacked barplot

```{r fig.width=8, fig.height=5}
plot_bar(
  ps_disease,
  fill = "Genus"
) +
  theme_bw() +
  labs(title = "Top 10 genera by disease group")
```

# Extract OTU table

```{r}
otu <- as.data.frame(
  otu_table(ps_rel)
)

if(taxa_are_rows(ps_rel)){
  otu <- t(otu)
}

otu <- as.data.frame(otu)
```

# Extract metadata

```{r}
meta <- sample_data(ps_rel) |>
  data.frame()

otu$Disease <- meta[
  rownames(otu),
  "Disease.MESH.ID"
]
```

# Differential abundance analysis

```{r}
results <- data.frame()

taxa <- setdiff(
  colnames(otu),
  "Disease"
)

for(tax in taxa){

  vals <- otu[[tax]]

  if(length(unique(vals)) < 2){
    next
  }

  if(sum(vals) == 0){
    next
  }

  test <- tryCatch(
    wilcox.test(vals ~ otu$Disease),
    error = function(e) NULL
  )

  if(!is.null(test)){

    group1 <- unique(otu$Disease)[1]
    group2 <- unique(otu$Disease)[2]

    median1 <- median(
      vals[otu$Disease == group1],
      na.rm = TRUE
    )

    median2 <- median(
      vals[otu$Disease == group2],
      na.rm = TRUE
    )

    log2FC <- log2(
      (median2 + 1e-6) /
      (median1 + 1e-6)
    )

    results <- rbind(
      results,
      data.frame(
        Taxon = tax,
        pvalue = test$p.value,
        log2FC = log2FC
      )
    )
  }
}

results$padj <- p.adjust(
  results$pvalue,
  method = "BH"
)

results <- results |>
  arrange(padj)

head(results)
```

# Volcano plot

```{r fig.width=7, fig.height=5}
results$logp <- -log10(results$padj)

ggplot(
  results,
  aes(x = log2FC,
      y = logp)
) +
  geom_point(alpha = 0.7) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed"
  ) +
  theme_bw() +
  labs(
    title = "Differential abundance",
    x = "log2 fold change",
    y = "-log10 adjusted p-value"
  )
```

# Heatmap of top taxa

```{r fig.width=8, fig.height=8}
top_taxa <- results |>
  filter(padj < 0.05) |>
  slice_head(n = 20) |>
  pull(Taxon)

if(length(top_taxa) < 20){

  top_taxa <- results |>
    slice_head(n = 20) |>
    pull(Taxon)
}

mat <- otu[, top_taxa]

mat_scaled <- scale(mat)

annotation <- data.frame(
  Disease = otu$Disease
)

rownames(annotation) <- rownames(otu)

pheatmap(
  t(mat_scaled),
  annotation_col = annotation,
  show_colnames = FALSE,
  fontsize_row = 8
)
```

# China-only subset analysis

```{r}
ps_china <- subset_samples(
  ps_rel,
  Project.ID == "PRJNA422434"
)
```

# China-only Bray-Curtis

```{r}
bray_china <- distance(
  ps_china,
  method = "bray"
)
```

# China-only ordination

```{r}
ord_china <- ordinate(
  ps_china,
  method = "PCoA",
  distance = bray_china
)
```

# China-only PCoA

```{r fig.width=7, fig.height=5}
plot_ordination(
  ps_china,
  ord_china,
  color = "Disease.MESH.ID"
) +
  geom_point(size = 3) +
  stat_ellipse() +
  theme_bw()
```

# China-only PERMANOVA

```{r}
sample_china <- sample_data(ps_china) |>
  data.frame()

adonis2(
  bray_china ~ Disease.MESH.ID,
  data = sample_china,
  permutations = 999
)
```

# CLR transformation

```{r}
ps_clr <- microbiome::transform(
  ps,
  "clr"
)
```

# Euclidean distance on CLR-transformed data

```{r}
clr_dist <- distance(
  ps_clr,
  method = "euclidean"
)
```

# CLR PCoA

```{r}
ord_clr <- ordinate(
  ps_clr,
  method = "PCoA",
  distance = clr_dist
)
```

# CLR ordination plot

```{r fig.width=8, fig.height=5}
plot_ordination(
  ps_clr,
  ord_clr,
  color = "Disease.MESH.ID",
  shape = "Project.ID"
) +
  geom_point(size = 3) +
  theme_bw()
```