---
title: "Efter_filtrering_alpha_beta"
output: html_document
---

# Sæt directory 

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Loading libraries
```{r}
library(phyloseq)
library(ggplot2)
library(vegan)
library(dplyr)
library(tibble)
```

# Load phyloseq  - rå data 
```{r}
load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
ps_genus <- pylsq
rm(pylsq) # fjerne pylsq, så man kan have mere end et. 

# Undersøger de to phylseq
ps_genus
```

# Kun kina 
```{r}
ps_china <- subset_samples(ps_genus,
                           Project.ID == "PRJNA422434")
```

# Analyse af alpha diversitet 
Calculation of alpha diversity  - Shannon 
The function estimate_richness() calculates alpha-diversity estimates for each sample, while plot_richness() provides quick visualisations of those estimates across samples or sample groups.

```{r}
# Calculate Shannon diversity for all samples
shannon <- estimate_richness(
  ps_china,
  measures = c("Shannon")
)

# The function estimate_richness() calculates the selected diversity measures for every sample in the phyloseq object and returns the results as a data frame.


head(shannon)
```
Visualising alpha diversity - Shannon 

The function plot_richness() creates quick exploratory visualisations of alpha-diversity patterns across samples. The simplest possible plot displays diversity values for each sample individually.

```{r}
# Basic alpha-diversity plot
plot_richness(
  ps_china,
  measures = c("Shannon")
) 
```
Shannon ud fra sygdom 
```{r}
# Group samples by disease 
plot_richness(
  ps_china,
  measures = c("Shannon"),
  x = "Disease.MESH.ID"
) +
  geom_boxplot(alpha = 0.6) +
  theme_bw()
```
# Statestik af alpha 
The code below calculates both Shannon diversity, converts the sample names into an explicit column (sample_id), extracts the sample metadata, and finally merges everything into a single table.


# forbered data 
```{r}
# Calculate Shannon diversity
shannon <- estimate_richness(
  ps_china,
  measures = "Shannon"
) %>%
  rownames_to_column("sample_id")


# Extract sample metadata
sample_df <- sample_data(ps_china) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

# Combine diversity estimates and metadata
alpha_diversity <- inner_join(
  shannon,
  sample_df,
  by = "sample_id"
) 

head(alpha_diversity)
```
Undersøg:
Do microbial communities from disese states show significantly different alpha diversity?
```{r}
# De forskellife disease status undersøges: 
table(alpha_diversity$Disease.MESH.ID)

# der er kun to så en wilcox.test bruges 
wilcox_shannon <- wilcox.test(
  Shannon ~ Disease.MESH.ID,
  data = alpha_diversity
)
wilcox_shannon

```
# Måske noget med højde for reads?? 


# Analyse af beta diversitet 

## Visualisering af beta-diversitet for genus 
# Plottet viser "Hvor ens eller forskellige samples er i deres mikrobielle sammensætning."
# Viser den omtrentlige spredning for hver gruppe.

```{r fig.width=8, fig.height=5}
# compositional-normalisering til sum = 1
ps_rel_c <- transform_sample_counts(ps_china, function(x) x / sum(x))
# beregning af Bray-dist 
bray_dist <- distance(ps_rel_c, method = "bray")
# Ordinate - Reducerer komplekse afstande til 2D-koordinater så man kan plotte samples.
ord_bray <- ordinate(ps_rel_c, method = "PCoA", distance = bray_dist)

#Plot - for at se hvor forskellige micorbiom samples er
plot_ordination(ps_rel_c, ord_bray, color = "Disease.MESH.ID") +
  geom_point(size = 2, alpha = 0.7) +
  stat_ellipse() +
  theme_bw() +
  labs(title = "Bray–Curtis PCoA coloured by Disease.MESH.ID genus")
```

# Undersøg den statistiske forskel mellem grupperne 

tester statistisk:
Om mikrobiomets sammensætning varierer med sygdom 
```{r adonis-adjusted}
# Omdannelse af meta data til en dataframe 
sample_df_2 <- sample_data(ps_rel_c) |> data.frame()


# Permanova med både project og disease 
adonis2(bray_dist ~ Disease.MESH.ID,
        data = sample_df_2,
        permutations = 999)
```

# Dispersion
However, PERMANOVA results should always be interpreted carefully because significant results can sometimes be driven not only by differences in group location, but also by differences in within-group dispersion.

Checking homogeneity of dispersion

To evaluate this, we can use betadisper() from the vegan package.
```{r}
dispersion_bray <- betadisper(
  bray_dist,
  sample_df_2$Disease.MESH.ID
)

anova(dispersion_bray)
plot(dispersion_bray)

```
If dispersion differs strongly between groups, then a significant PERMANOVA result should be interpreted with caution.

# Resultat 
F = 1.4089
p = 0.236

Det betyder: XXXXX




