---
title: "Emilie"
output: html_document
---

Først instaleres phyloseq
```{r}
# if (!require("BiocManager", quietly = TRUE))
    # install.packages("BiocManager")

# BiocManager::install("phyloseq")
```

```{r}
# install.packages("lifecycle")
```

```{r}
# if (!require("BiocManager", quietly = TRUE))
   #  install.packages("BiocManager")

# BiocManager::install("microbiome")
```

Load the required R packages
The Rmd begins by loading the packages used for microbiome data handling, data wrangling, plotting, model fitting, and ordination:
This hidden setup chunk tells R Markdown to show the code in the rendered practical while suppressing routine package messages and warnings. That keeps the rendered HTML focused on the analysis output rather than package startup text.
```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(ggplot2)
library(tidyverse)
library(ggh4x)
library(lme4)
library(lmerTest)
library(broom.mixed)
library(microbiome)
```

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

Load the phyloseq object - genus 
```{r}
# Read the saved phyloseq  - genus 
obj_name <- load("project/data/D003924_Diabetes_Mellitus,_T/genus/phyloseq.RData")
# get() tager et navn som tekst og finder det rigtige objekt i R-miljøet.
pylsq <- get(obj_name)

# Print a compact summary of the phyloseq object, including sample and taxon counts.

# Inspect the sample metadata table attached to the phyloseq object.
sample_data(pylsq)

# Antal taxa
ntaxa(pylsq)
# 58

# List the available taxonomic ranks, such as kingdom, phylum, genus, and species.
rank_names(pylsq)
# "Domain"  "Phylum"  "Class"   "Order"   "Family"  "Genus"   "Species"

# List the metadata variables that can be used for filtering, plotting, or modelling.
sample_variables(pylsq)


# Taxa-table 
tax_table(pylsq)
head(as.data.frame(tax_table(pylsq)))

```

Calculation of alpha diversity  - Shannon 
The function estimate_richness() calculates alpha-diversity estimates for each sample, while plot_richness() provides quick visualisations of those estimates across samples or sample groups.

```{r}
# Calculate Shannon diversity for all samples
shannon <- estimate_richness(
  pylsq,
  measures = c("Shannon")
)

# The function estimate_richness() calculates the selected diversity measures for every sample in the phyloseq object and returns the results as a data frame.


head(shannon)
```

Visualising alpha diversity - Shannon 

The function plot_richness() creates quick exploratory visualisations of alpha-diversity patterns across samples. The simplest possible plot displays diversity values for each sample individually.

```{r}
# Basic alpha-diversity plot
plot_richness(
  pylsq,
  measures = c("Shannon")
) 
```


Gruped data
```{r}
# Group samples by disease 
plot_richness(
  pylsq,
  measures = c("Shannon"),
  x = "Disease.MESH.ID"
) +
  geom_boxplot(alpha = 0.6) +
  theme_bw()
```

```{r}
# Colour samples according to another metadata variable
pylsq_noNA <- subset_samples(pylsq, !is.na(BMI))

plot_richness(
  pylsq_noNA,
  measures = c("Shannon"),
  x = "Disease.MESH.ID",
  color = "BMI"
)
```

```{r}
# Colour samples according to another metadata variable
plot_richness(
  pylsq,
  measures = c("Shannon"),
  x = "Disease.MESH.ID",
  color = "Host.age"
)
```

The code below calculates both Shannon diversity and Observed richness separately, converts the sample names into an explicit column (sample_id), extracts the sample metadata, and finally merges everything into a single table.


```{r}
# Calculate Shannon diversity
shannon <- estimate_richness(
  pylsq,
  measures = "Shannon"
) %>%
  rownames_to_column("sample_id")


# Extract sample metadata
sample_df <- sample_data(pylsq) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

# Combine diversity estimates and metadata
alpha_diversity <- inner_join(
  shannon,
  sample_df,
  by = "sample_id"
) 

head(alpha_diversity)
```

The resulting alpha_diversity data frame contains:

one row per sample,
alpha-diversity estimates,
and all associated sample metadata variables.
This format is particularly useful because it allows us to use standard tidyverse and statistical functions for downstream analyses.

Undersøg:
Do microbial communities from disese states show significantly different alpha diversity?
```{r}
# De forskellife disease status undersøges: 
table(alpha_diversity$Disease.MESH.ID)

# der er kun to så en wilcox.test bruges 
wilcox_shannon <- wilcox.test(
  Shannon ~ Disease.MESH.ID,
  data = alpha_diversity
)
wilcox_shannon

```


Undersøg:
Do microbial communities from sex show significantly different alpha diversity?
```{r}
# De forskellife disease status undersøges: 
table(alpha_diversity$Sex)

alpha_diversity_noNA <- subset(
  alpha_diversity,
  !is.na(Sex) & Sex != ""
)
# der er kun to så en wilcox.test bruges 
wilcox_shannon_2 <- wilcox.test(
  Shannon ~ Sex,
  data = alpha_diversity_noNA
)

wilcox_shannon_2
```
Undersøg:
Do microbial communities from different contries show significantly different alpha diversity?
```{r}
# De forskellife lande  undersøges: 
table(alpha_diversity$Country)

# der er mere end to  
kruskal_country <- kruskal.test(
  Shannon ~ Country,
  data = alpha_diversity
)
kruskal_country 

```

Visualise alpha diversity 

```{r}
ggplot(alpha_diversity, aes(x = Disease.MESH.ID, y = Shannon)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  theme_bw()
```

We can explore whether samples with more reads also tend to have higher diversity estimates.

```{r}
# Plot relationship
ggplot(alpha_diversity, aes(x = Nr..reads.sequenced, y = Shannon)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

Test it statistically 
```{r}
cor.test(
  alpha_diversity$Nr..reads.sequenced,
  alpha_diversity$Shannon,
  method = "spearman"
)
```
Er der stadig forskel på syg og rask, når der tages højde for reads

```{r}
model <- lm(
  Shannon ~ Disease.MESH.ID + Nr..reads.sequenced,
  data = alpha_diversity
)

summary(model)

ggplot(alpha_diversity,
       aes(x = Nr..reads.sequenced,
           y = Shannon,
           color = Disease.MESH.ID)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

"No significant difference in Shannon alpha diversity was observed between disease groups after adjusting for sequencing depth (linear model, p = 0.829). Sequencing depth was strongly associated with Shannon diversity (p < 0.001)."

log transformation af reads 
```{r}
model2 <- lm(
  Shannon ~ Disease.MESH.ID + log10(Nr..reads.sequenced),
  data = alpha_diversity
)

summary(model2)

ggplot(alpha_diversity,
       aes(x = log10(Nr..reads.sequenced),
           y = Shannon,
           color = Disease.MESH.ID)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()
```

Undersøg forskel mellem reads og nr.reads.seq...
```{r}
alpha_diversity$reads_phyloseq <- sample_sums(pylsq)

cor(
  alpha_diversity$reads_phyloseq,
  alpha_diversity$Nr..reads.sequenced
)

summary(sample_sums(pylsq))

otu_table(pylsq)[1:10,1:10]
```

Estimation and visualisation of beta diversity

Presence/absence transformation

In some cases, we may want to focus only on whether a taxon is detected or not, without considering its abundance.

```{r}
ps_pa <- transform_sample_counts(
  pylsq,
  function(x){ifelse(x > 0, 1, 0)}
)

sample_sums(ps_pa)

otu_table(ps_pa)[1:5,1:5]
```

Step 2: Estimating beta-diversity distances

The result is a distance matrix in which:

small values indicate that two samples are compositionally similar,
and large values indicate that they are more different.


In this session, we will use two common beta-diversity metrics:

Bray-Curtis distance, which is based on abundance differences between samples,
and Jaccard distance, which focuses on shared versus unshared taxa.

Bray-Curtis distance

Bray-Curtis distance is one of the most widely used measures in microbial community analysis. It is sensitive to differences in taxon abundances and is therefore commonly applied to relative-abundance data.
```{r}
bray_dist <- distance(
  pylsq,
  method = "bray"
)

bray_dist
```

Jaccard distance

Jaccard distance is often applied to presence/absence data and focuses on whether the same taxa are present in two samples, regardless of their abundances.
```{r}
jaccard_dist <- distance(
  ps_pa,
  method = "jaccard"
)

jaccard_dist
```

 Step 3: Ordination with phyloseq
 phyloseq provides the function ordinate() to compute ordinations and plot_ordination() to visualise them.
 
 PCoA based on Bray-Curtis distance

PCoA is a useful first ordination method because it attempts to represent the pairwise sample distances in a lower-dimensional Euclidean space.

```{r}
ord_bray <- ordinate(
  pylsq,
  method = "PCoA",
  distance = bray_dist
)

plot_ordination(
  pylsq,
  ord_bray,
  type = "samples",
  color = "Disease.MESH.ID"
)
```

Extracting the ordination coordinates

The built-in plot is useful for quick inspection, but it is often convenient to extract the ordination coordinates as a standard data frame so that we can customise the visualisation further with ggplot2.

```{r}
ord_bray_df <- plot_ordination(
  pylsq,
  ord_bray,
  type = "samples",
  justDF = TRUE
)

ord_bray_df$sample_id <- rownames(ord_bray_df)

sample_df <- sample_data(pylsq) %>%
  data.frame() %>%
  rownames_to_column("sample_id")

ord_bray_df <- inner_join(
  ord_bray_df,
  sample_df,
  by = "sample_id"
)

head(ord_bray_df)

ggplot(ord_bray_df, aes(x = Axis.1, y = Axis.2, color = Disease.MESH.ID.x)) +
  geom_point(size = 3) +
  theme_bw()
```

NMDS based on Bray-Curtis distance

NMDS is another commonly used ordination method in ecology and microbiome analysis. Unlike PCoA, it focuses on preserving the rank order of distances rather than the distances themselves.

```{r}
ord_nmds <- ordinate(
  pylsq,
  method = "NMDS",
  distance = "bray"
)

plot_ordination(
  pylsq,
  ord_nmds,
  type = "samples",
  color = "Disease.MESH.ID"
)
```
Step 4: Statistical testing of beta diversity

Combining the distance matrix with sample metadata

Before running the test, we first extract the sample metadata associated with the phyloseq object.
```{r}
library(vegan)

sample_df <- sample_data(pylsq) %>%
  data.frame()

head(sample_df)
```

Testing whether disease state differ in composition
```{r}
adonis_bray <- adonis2(
  bray_dist ~ Disease.MESH.ID,
  data = sample_df,
  permutations = 999
)

adonis_bray
```
Der er en significant forskel (0,001), men effeitstørrelsen er lille 0.01411 --> sygdom forklarer kun ca. 1.4% af variationen i microbiomet

However, PERMANOVA results should always be interpreted carefully because significant results can sometimes be driven not only by differences in group location, but also by differences in within-group dispersion.

Checking homogeneity of dispersion

To evaluate this, we can use betadisper() from the vegan package.
```{r}
dispersion_bray <- betadisper(
  bray_dist,
  sample_df$Disease.MESH.ID
)

anova(dispersion_bray)
plot(dispersion_bray)
```
If dispersion differs strongly between groups, then a significant PERMANOVA result should be interpreted with caution.

Hvis den IKKE er signifikant:
→ din PERMANOVA er mere troværdig.

Hvis den ER signifikant:
→ forskellen kan skyldes forskellig spredning snarere end forskellig community composition.

Eftersom den er meget signifikant kan forskellen fra adonis_bray spredning snarere end forskellig community composition.



Gruped data - country 

```{r}
# Group samples by disease 
plot_richness(
  pylsq,
  measures = c("Shannon"),
  x = "Project.ID"
) +
  geom_boxplot(alpha = 0.6) +
  theme_bw()

# Stat 
kruskal_country <- kruskal.test(
  Shannon ~ Project.ID,
  data = alpha_diversity
)
kruskal_country 

# pair
pw_shannon <- pairwise.wilcox.test(
  x = alpha_diversity$Shannon,
  g = alpha_diversity$Project.ID,
  p.adjust.method = "BH"
)
pw_shannon
```






