---
title: "Final_analysis"
output: html_document
---
```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```


```{r}
library(phyloseq)
library(vegan)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(ggplot2)
library(ggh4x)
library(broom)
library(microbiome)
```


## Load and filter species data

```{r load-data}
load("project/data/D003924_Diabetes_Mellitus,_T/species/D003924_phyloseq_updated_species.RData")
ps <- subset_samples(pylsq, Project.ID == "PRJNA422434")
rm(pylsq)

ps
table(sample_data(ps)$Disease.MESH.ID)
ntaxa(ps)
```




MetaPhlAn kunne ikke identificere organismen sikkert helt ned til species-niveau.
Den ved Family = Lachnospiraceae med høj sikkerhed. Men den ved ikke præcis hvilken genus eller species. Denne bakterie ligger sandsynligvis et sted i Lachnospiraceae-familien, og den nærmeste reference vi fandt var Roseburia.
Den siger s_Roseburia og ikke  intestinalis
Roseburia hominis
Roseburia faecis det viser at men højere til familien jensen men mangler et fornanvn species niveau 




```{r}
#longformat
ps_rel  <- transform_sample_counts(ps, function(x) x / sum(x))
otu_rel <- as(otu_table(ps_rel), "matrix") %>% t() %>% as.data.frame()
meta_rel <- data.frame(sample_data(ps_rel))
tax_rel  <- as(tax_table(ps_rel), "matrix") %>% as.data.frame()

plot_bar(ps_rel, fill = "Species") +
  geom_bar(aes(fill = Species), stat = "identity", colour = NA) +
  facet_nested(~ Sex + Disease.MESH.ID, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        legend.position = "none") +
  labs(title = "Genus-level composition by Sex × Disease",
       fill = "Genus")
```

```{r}
#DaTA LONG FORMAT
dat_rel <- otu_rel %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id, names_to = "taxon", values_to = "rel_abund") %>%
  left_join(meta_rel %>% rownames_to_column("sample_id"), by = "sample_id") %>%
  left_join(tax_rel  %>% rownames_to_column("taxon"),     by = "taxon")

```





```{r}
one_taxon <- unique(dat_rel$taxon)[3]
# one_taxon <- unique(dat_rel$taxon)[5]
# one_taxon <- unique(dat_rel$taxon)[20]

# Comparison between disease groups
dat_rel %>%
  filter(taxon == one_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    theme_bw() +
    labs(title = paste("Taxon", one_taxon, "by disease"))

```
```{r}
# Comparison between sexes within T2D
dat_rel %>%
  filter(taxon == one_taxon, Disease.MESH.ID == "D003924") %>%
  ggplot(aes(x = Sex, y = rel_abund, fill = Sex)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    theme_bw() +
    labs(title = paste("Taxon", one_taxon, "by sex (T2D only)"))
```
```{r}
# Zero inflation plot
dat_rel %>%
  group_by(taxon, Disease.MESH.ID) %>%
  summarise(
    prop_zero = mean(rel_abund == 0),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = prop_zero, fill = Disease.MESH.ID)) +
  geom_histogram(
    binwidth = 0.05,
    colour = "white",
    alpha = 0.7
  ) +
  scale_fill_manual(
    values = c(
      "D003924" = "#E69F00",
      "D006262" = "#56B4E9"
    )
  ) +
  facet_wrap(~Disease.MESH.ID) +
  labs(
    x = "Proportion of samples where relative abundance = 0",
    y = "Number of taxa",
    title = "Zero inflation by disease group"
  ) +
  theme_bw() +
  theme(legend.position = "none")

```
```{r}
#Right-skewed distributions



set.seed(42)

example_taxa <- dat_rel %>%
  filter(Disease.MESH.ID == "D003924") %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    prop_nonzero = mean(rel_abund > 0),
    .groups = "drop"
  ) %>%
  filter(
    prop_nonzero > 0.4,
    prop_nonzero < 0.9
  ) %>%
  slice_sample(n = 6) %>%
  pull(taxon)


dat_rel %>%
  filter(
    Disease.MESH.ID == "D003924",
    taxon %in% example_taxa,
    rel_abund > 0
  ) %>%
  mutate(label = Species) %>%
  ggplot(aes(x = rel_abund)) +
  geom_histogram(
    bins = 20,
    fill = "steelblue",
    colour = "white"
  ) +
  facet_wrap(
    ~label,
    scales = "free"
  ) +
  labs(
    x = "Relative abundance (non-zero values only)",
    y = "Count",
    title = "Right-skewed distributions in selected taxa (T2D)"
  ) +
  theme_bw()

```
```{r}
dat_rel %>%
  filter(
    Disease.MESH.ID == "D003924",
    taxon %in% example_taxa,
    rel_abund > 0
  ) %>%
  mutate(label = Species) %>%
  ggplot(aes(sample = rel_abund)) +
  stat_qq() +
  stat_qq_line(colour = "red") +
  facet_wrap(
    ~label,
    scales = "free"
  ) +
  labs(
    title = "Q-Q plots of selected taxa"
  ) +
  theme_bw()
```
```{r}
wilcox_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    p_value = wilcox.test(rel_abund ~ Disease.MESH.ID, exact = FALSE)$p.value,
    mean_T2D = mean(rel_abund[Disease.MESH.ID == "D003924"]),
    mean_Healthy = mean(rel_abund[Disease.MESH.ID == "D006262"]),
    log2_fc = log2((mean_T2D + 1e-6) / (mean_Healthy + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(
    q_value = p.adjust(p_value, method = "BH"),
    direction = ifelse(log2_fc > 0, "T2D-enriched", "Healthy-enriched")
  )


```


```{r}
# Step 1 — Run one Wilcoxon test per taxon
wilcox_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  summarise(
    p_value      = wilcox.test(rel_abund ~ Disease.MESH.ID, exact = FALSE)$p.value,
    mean_T2D     = mean(rel_abund[Disease.MESH.ID == "D003924"]),
    mean_Healthy = mean(rel_abund[Disease.MESH.ID == "D006262"]),
    log2_fc      = log2((mean_T2D + 1e-6) / (mean_Healthy + 1e-6)),
    .groups = "drop"
  ) %>%
  mutate(q_value = p.adjust(p_value, method = "BH"))

head(wilcox_results, 10)

```
```{r}
top_taxon <- wilcox_T2D_depleted %>% slice(1) %>% pull(taxon)
top_label <- wilcox_T2D_depleted %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-depleted (Wilcoxon):", top_label))
```


```{r}
### Top T2D-enriched (Wilcoxon)
wilcox_T2D_enriched <- wilcox_results %>%
  select(taxon, Genus, Species, mean_T2D, mean_Healthy, log2_fc, p_value, q_value) %>%
  arrange(-log2_fc) %>%
  head(10)

wilcox_T2D_enriched




```
```{r}
wilcox_Healthy_enriched <- wilcox_results %>%
  select(taxon, Genus, Species, mean_T2D, mean_Healthy, log2_fc, p_value, q_value, direction) %>%
  arrange(log2_fc) %>%
  head(10)

wilcox_Healthy_enriched
```



```{r}
top_taxon <- wilcox_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- wilcox_T2D_enriched %>% slice(1) %>% pull(Species)


top_taxon <- wilcox_T2D_enriched %>% slice(1) %>% pull(taxon)
top_label <- wilcox_T2D_enriched %>% slice(1) %>% pull(Species)

dat_rel %>%
  filter(taxon == top_taxon) %>%
  ggplot(aes(x = Disease.MESH.ID, y = rel_abund, fill = Disease.MESH.ID)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.4) +
    scale_fill_manual(values = c("D003924" = "#E69F00", "D006262" = "#56B4E9")) +
    theme_bw() +
    labs(title = paste("Top T2D-enriched (Wilcoxon):", top_label))




```

### Volcano plot — Wilcoxon
```{r}
wilcox_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & log2_fc > 0 ~ "T2D-enriched",
      q_value < 0.05 & log2_fc < 0 ~ "T2D-depleted",
      TRUE                         ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = log2_fc, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "log2 fold change (T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "Wilcoxon: species-level disease effect") +
    theme_bw()



```
```{r}

### Top T2D-enriched (GLM unadjusted)
glm_T2D_enriched <- glm_disease_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy,
         std.error, statistic, p.value, q_value) %>%
  arrange(-fold_change_T2D_vs_Healthy) %>%
  head(10)

glm_T2D_enriched


```



### Volcano plot — GLM unadjusted

```{r glm-unadj-volcano, fig.width=8, fig.height=5}
glm_disease_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "T2D-depleted",
      TRUE                          ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
    geom_point(alpha = 0.6, size = 2) +
    scale_color_manual(values = c("T2D-enriched" = "#E74C3C",
                                  "T2D-depleted" = "#3498DB",
                                  "Not significant" = "grey60")) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
    labs(x = "GLM estimate (log10 RA, T2D vs Healthy)",
         y = "-log10(q-value)", color = NULL,
         title = "GLM unadjusted: species-level disease effect") +
    theme_bw()

```


```{r}



#Mest Healthy-enriched (størst negativ fold change)

wilcox_Healthy_enriched <- wilcox_results %>%
  select(taxon, Genus, Species,
         mean_T2D, mean_Healthy,
         log2_fc, p_value, q_value) %>%
  arrange(log2_fc)

head(wilcox_Healthy_enriched, 10)

```

## Primary adjusted model — Sex + Host.age (n = 359)

```{r glm-adj-results}
glm_disease_adjusted_sex_results <- dat_rel %>%
  mutate(
    Disease.MESH.ID = factor(Disease.MESH.ID, levels = c("D006262", "D003924")),
    Sex             = factor(Sex,             levels = c("Female", "Male")),
    Host.age        = as.numeric(Host.age)
  ) %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(log10(rel_abund + 1e-6) ~ Disease.MESH.ID + Sex + Host.age,
               data = .x, family = gaussian())
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease.MESH.IDD003924") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_T2D_vs_Healthy_adj = 10^estimate
  )

```

> *Taught in: W21 reference Rmd §7 "Adjust the linear model" (same GLM with covariates; teacher uses `status`, here `Sex` + `Host.age`). Covariate adjustment: Alberdi reading §5. Sex/Age are the only complete, polymorphic metadata variables in this cohort.*

### Top T2D-depleted (GLM adj. Sex + Age)

```{r glm-adj-depleted}
glm_adj_T2D_depleted <- glm_disease_adjusted_sex_results %>%
  select(taxon, Genus, Species, estimate, fold_change_T2D_vs_Healthy_adj,
         std.error, statistic, p.value, q_value) %>%
  arrange(fold_change_T2D_vs_Healthy_adj) %>%
  head(10)

glm_adj_T2D_depleted
```

```{r}
# --------------------------------------------------
# GLM unadjusted: T2D vs Healthy
# --------------------------------------------------
dat_rel <- dat_rel %>%
  mutate(
    Disease_group = factor(
      Disease.MESH.ID,
      levels = c("D006262", "D003924"),
      labels = c("Healthy", "T2D")
    )
  )


glm_disease_results <- dat_rel %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(
      log10(rel_abund + 1e-6) ~ Disease_group,
      data = .x
    )
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease_groupT2D") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    fold_change_T2D_vs_Healthy = 10^estimate,
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  )

# --------------------------------------------------
# Top Healthy-enriched taxa in GLM
# --------------------------------------------------

glm_Healthy_enriched <- glm_disease_results %>%
  select(
    taxon, Genus, Species,
    estimate,
    fold_change_T2D_vs_Healthy,
    std.error, statistic, p.value, q_value,
    direction
  ) %>%
  arrange(fold_change_T2D_vs_Healthy) %>%
  head(10)

glm_Healthy_enriched

# Plot strongest Healthy-enriched GLM hit
dat_rel %>%
  filter(taxon == glm_Healthy_enriched %>% slice(1) %>% pull(taxon)) %>%
  ggplot(aes(x = Disease_group, y = rel_abund, fill = Disease_group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_fill_manual(values = c(
    "Healthy" = "#3498DB",
    "T2D" = "#E74C3C"
  )) +
  labs(
    x = "",
    y = "Relative abundance",
    title = paste("Top Healthy-enriched GLM hit:",
                  glm_Healthy_enriched %>% slice(1) %>% pull(Species))
  ) +
  theme_bw()

# --------------------------------------------------
# Top T2D-enriched taxa in GLM
# --------------------------------------------------

glm_T2D_enriched <- glm_disease_results %>%
  select(
    taxon, Genus, Species,
    estimate,
    fold_change_T2D_vs_Healthy,
    std.error, statistic, p.value, q_value,
    direction
  ) %>%
  arrange(desc(fold_change_T2D_vs_Healthy)) %>%
  head(10)

glm_T2D_enriched

# Plot strongest T2D-enriched GLM hit
dat_rel %>%
  filter(taxon == glm_T2D_enriched %>% slice(1) %>% pull(taxon)) %>%
  ggplot(aes(x = Disease_group, y = rel_abund, fill = Disease_group)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  scale_fill_manual(values = c(
    "Healthy" = "#3498DB",
    "T2D" = "#E74C3C"
  )) +
  labs(
    x = "",
    y = "Relative abundance",
    title = paste("Top T2D-enriched GLM hit:",
                  glm_T2D_enriched %>% slice(1) %>% pull(Species))
  ) +
  theme_bw()

# --------------------------------------------------
# Volcano plot: GLM unadjusted
# --------------------------------------------------

glm_disease_results %>%
  mutate(
    neg_log10_q = -log10(q_value),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = estimate, y = neg_log10_q, color = direction)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c(
    "T2D-enriched" = "#E74C3C",
    "Healthy-enriched" = "#3498DB",
    "Not significant" = "grey60"
  )) +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    x = "GLM estimate: log10 relative abundance (T2D vs Healthy)",
    y = "-log10(q-value)",
    color = NULL,
    title = "GLM: T2D effect in gut microbiome"
  ) +
  theme_bw()

# --------------------------------------------------
# Bar chart of top significant taxa, GLM
# --------------------------------------------------

bind_rows(
  glm_disease_results %>%
    filter(q_value < 0.05) %>%
    arrange(estimate) %>%
    head(10),

  glm_disease_results %>%
    filter(q_value < 0.05) %>%
    arrange(desc(estimate)) %>%
    head(10)
) %>%
  distinct() %>%
  mutate(
    label = Species,
    direction = ifelse(
      estimate > 0,
      "T2D-enriched",
      "Healthy-enriched"
    )
  ) %>%
  ggplot(aes(
    x = reorder(label, -q_value),
    y = estimate,
    fill = direction
  )) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "T2D-enriched" = "#E74C3C",
    "Healthy-enriched" = "#3498DB"
  )) +
  labs(
    x = NULL,
    y = "GLM estimate: log10 relative abundance",
    fill = NULL,
    title = "Top differentially abundant taxa (GLM)"
  ) +
  theme_bw()

# --------------------------------------------------
# Compare GLM and Wilcoxon results
# --------------------------------------------------

model_comparison <- wilcox_results %>%
  select(
    taxon, Genus, Species,
    wilcox_log2_fc = log2_fc,
    wilcox_q = q_value
  ) %>%
  left_join(
    glm_disease_results %>%
      select(
        taxon,
        glm_estimate = estimate,
        glm_fold_change = fold_change_T2D_vs_Healthy,
        glm_q = q_value
      ),
    by = "taxon"
  ) %>%
  mutate(
    wilcox_direction = case_when(
      wilcox_log2_fc > 0 ~ "T2D-enriched",
      wilcox_log2_fc < 0 ~ "Healthy-enriched",
      TRUE ~ "No direction"
    ),
    glm_direction = case_when(
      glm_estimate > 0 ~ "T2D-enriched",
      glm_estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "No direction"
    ),
    same_direction = wilcox_direction == glm_direction
  )

model_comparison %>%
  arrange(glm_q) %>%
  head(20)
```
```{r}

# --------------------------------------------------
# CLR transformation
# --------------------------------------------------

ps_clr <- transform_sample_counts(ps, function(x) {
  x <- x + 1e-9
  log(x) - mean(log(x))
})

otu_clr <- as(otu_table(ps_clr), "matrix") %>%
  t() %>%
  as.data.frame()

meta_clr <- as(sample_data(ps_clr), "matrix") %>%
  as.data.frame()

tax_clr <- as(tax_table(ps_clr), "matrix") %>%
  as.data.frame()

dat_clr <- otu_clr %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(
    -sample_id,
    names_to = "taxon",
    values_to = "clr"
  ) %>%
  left_join(
    meta_clr %>%
      as("data.frame") %>%
      rownames_to_column("sample_id"),
    by = "sample_id"
  ) %>%
  left_join(
    tax_clr %>%
      rownames_to_column("taxon"),
    by = "taxon"
  ) %>%
  mutate(
    Disease_group = factor(
      Disease.MESH.ID,
      levels = c("D006262", "D003924"),
      labels = c("Healthy", "T2D")
    )
  )


```



```{r}
#CLR model uden justering
clr_results <- dat_clr %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(clr ~ Disease_group, data = .x)
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease_groupT2D") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  )

clr_results %>%
  arrange(q_value) %>%
  head(20)
```




#Adjusted CLR model

```{r}
clr_adj_results <- dat_clr %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(
      clr ~ Disease_group + Sex + BMI + Host.age,
      data = .x
    )
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease_groupT2D") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  )

clr_adj_results %>%
  arrange(q_value) %>%
  head(20)

```

```{r}

#sex +dieases  clr
clr_adj_results <- dat_clr %>%
  group_by(taxon, Genus, Species) %>%
  group_modify(~ {
    fit <- glm(
      clr ~ Disease_group + Sex,
      data = .x
    )
    broom::tidy(fit)
  }) %>%
  ungroup() %>%
  filter(term == "Disease_groupT2D") %>%
  mutate(
    q_value = p.adjust(p.value, method = "BH"),
    direction = case_when(
      q_value < 0.05 & estimate > 0 ~ "T2D-enriched",
      q_value < 0.05 & estimate < 0 ~ "Healthy-enriched",
      TRUE ~ "Not significant"
    )
  )

clr_adj_results
```


PCA af CLR-data (T2D vs Healthy)

```{r}
# Community matrix
gut_clr_mat <- dat_clr %>%
  select(sample_id, taxon, clr) %>%
  pivot_wider(names_from = taxon, values_from = clr) %>%
  column_to_rownames("sample_id") %>%
  as.matrix()

gut_meta <- dat_clr %>%
  distinct(sample_id, Disease_group, Sex, BMI) %>%
  column_to_rownames("sample_id")

# PCA
gut_pca <- prcomp(gut_clr_mat,
                  center = TRUE,
                  scale. = FALSE)

pca_scores <- as.data.frame(gut_pca$x[,1:2]) %>%
  rownames_to_column("sample_id") %>%
  left_join(
    gut_meta %>%
      rownames_to_column("sample_id"),
    by = "sample_id"
  )

pca_var <- round(
  100 * gut_pca$sdev^2 /
    sum(gut_pca$sdev^2),
  1
)

ggplot(
  pca_scores,
  aes(
    PC1,
    PC2,
    color = Disease_group
  )
) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(
    values = c(
      "Healthy" = "#3498DB",
      "T2D" = "#E74C3C"
    )
  ) +
  labs(
    x = paste0("PC1 (", pca_var[1], "%)"),
    y = paste0("PC2 (", pca_var[2], "%)"),
    title = "PCA of CLR-transformed microbiome"
  ) +
  theme_bw()
```
RDA kan T2D foorklare vvariationen? 

NOVA DISEASE 
```{r}


rda_disease <- rda(
  gut_clr_mat ~ Disease_group,
  data = gut_meta
  
)
anova(
  rda_disease,
  permutations = 999
)

```

Pr(>F) = 0.01

så: T2D forklarer signifikant variation i mikrobiomet.


4. Test sygdomseffekten alene
```{r}
anova(
  rda_disease,
  by = "term",
  permutations = 999
)

```
Justeret model

```{r}
taxon_scores <- scores(
  rda_disease,
  display = "species"
)[,1]

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```


```{r}
taxon_scores <- scores(
  rda_disease,
  display = "species"
)[,1]

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))
```


```{r}
rda_taxa %>%
  head(20) %>%
  left_join(
    tax_clr %>%
      rownames_to_column("taxon"),
    by = "taxon"
  ) %>%
  mutate(
    direction =
      ifelse(
        rda1_loading > 0,
        "T2D-associated",
        "Healthy-associated"
      )
  ) %>%
  ggplot(
    aes(
      reorder(Species, rda1_loading),
      rda1_loading,
      fill = direction
    )
  ) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(
    values = c(
      "T2D-associated" = "#E74C3C",
      "Healthy-associated" = "#3498DB"
    )
  ) +
  theme_bw() +
  labs(
    x = NULL,
    y = "RDA1 loading",
    title = "Taxa driving T2D vs Healthy separation"
  )
```
```{r}
# --------------------------------------------------
# RDA biplot: T2D vs Healthy
# --------------------------------------------------

rda_disease <- rda(
  gut_clr_mat ~ Disease_group + Sex,
  data = gut_meta
)

anova(rda_disease, permutations = 999)
anova(rda_disease, by = "term", permutations = 999)

# Percent variance explained by RDA axes
eig <- rda_disease$CCA$eig
pct1 <- round(eig[1] / rda_disease$tot.chi * 100, 1)
pct2 <- round(eig[2] / rda_disease$tot.chi * 100, 1)

# Orient RDA1 so positive side points toward T2D
rda_site_scores_raw <- scores(rda_disease, display = "sites")

rda1_sign <- ifelse(
  mean(rda_site_scores_raw[gut_meta$Disease_group == "T2D", "RDA1"], na.rm = TRUE) >=
    mean(rda_site_scores_raw[gut_meta$Disease_group == "Healthy", "RDA1"], na.rm = TRUE),
  1,
  -1
)

# Site scores = sample positions
site_scores <- scores(rda_disease, display = "sites") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("sample_id") %>%
  left_join(
    gut_meta %>% rownames_to_column("sample_id"),
    by = "sample_id"
  )

# Biplot scores = arrows for explanatory variables
biplot_scores <- scores(rda_disease, display = "bp") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("variable")

# Taxon scores
taxon_scores <- scores(rda_disease, display = "species")[, 1] * rda1_sign

rda_taxa <- tibble(
  taxon = names(taxon_scores),
  rda1_loading = taxon_scores
) %>%
  arrange(desc(abs(rda1_loading)))

top_species_scores <- scores(rda_disease, display = "species") %>%
  as.data.frame() %>%
  mutate(RDA1 = RDA1 * rda1_sign) %>%
  rownames_to_column("taxon") %>%
  inner_join(rda_taxa %>% head(10), by = "taxon") %>%
  left_join(
    tax_clr %>% rownames_to_column("taxon"),
    by = "taxon"
  ) %>%
  mutate(label = Species)

# Scale arrows for visualisation
site_range <- max(abs(c(site_scores$RDA1, site_scores$RDA2)))

arrow_scale <- site_range * 0.8 /
  max(sqrt(biplot_scores$RDA1^2 + biplot_scores$RDA2^2))

species_scale <- site_range * 0.6 /
  max(sqrt(top_species_scores$RDA1^2 + top_species_scores$RDA2^2))

biplot_scores <- biplot_scores %>%
  mutate(
    RDA1 = RDA1 * arrow_scale,
    RDA2 = RDA2 * arrow_scale
  )

top_species_scores <- top_species_scores %>%
  mutate(
    RDA1 = RDA1 * species_scale,
    RDA2 = RDA2 * species_scale
  )

# Plot RDA biplot
ggplot() +
  geom_point(
    data = site_scores,
    aes(x = RDA1, y = RDA2, color = Disease_group, shape = Sex),
    size = 3,
    alpha = 0.85
  ) +
  geom_segment(
    data = top_species_scores,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.15, "cm")),
    color = "grey70",
    linewidth = 0.5
  ) +
  geom_text(
    data = top_species_scores,
    aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = label),
    size = 2.5,
    color = "grey50"
  ) +
  geom_segment(
    data = biplot_scores,
    aes(x = 0, y = 0, xend = RDA1, yend = RDA2),
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black",
    linewidth = 1
  ) +
  geom_text(
    data = biplot_scores,
    aes(x = RDA1 * 1.15, y = RDA2 * 1.15, label = variable),
    size = 3.5,
    fontface = "bold",
    color = "black"
  ) +
  scale_color_manual(values = c(
    "Healthy" = "#3498DB",
    "T2D" = "#E74C3C"
  )) +
  labs(
    x = paste0("RDA1 (", pct1, "% of total variance)"),
    y = paste0("RDA2 (", pct2, "% of total variance)"),
    color = "Disease group",
    shape = "Sex",
    title = "RDA biplot: CLR microbiome constrained by T2D status and sex"
  ) +
  theme_bw()
```

```{r}
rda_taxa %>%
  head(20) %>%
  left_join(
    tax_clr %>% rownames_to_column("taxon"),
    by = "taxon"
  ) %>%
  mutate(
    label = Species,
    direction = ifelse(
      rda1_loading > 0,
      "T2D-associated",
      "Healthy-associated"
    )
  ) %>%
  ggplot(aes(
    x = reorder(label, rda1_loading),
    y = rda1_loading,
    fill = direction
  )) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c(
    "T2D-associated" = "#E74C3C",
    "Healthy-associated" = "#3498DB"
  )) +
  labs(
    x = NULL,
    y = "RDA1 loading",
    fill = NULL,
    title = "Top 20 taxa driving T2D-associated RDA axis"
  ) +
  theme_bw()
```

